---
title: "RegCompassR 2.1.0 outer-OOF condition workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RegCompassR 2.1.0 outer-OOF condition workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

> Each broad cell type is trained, validated, and refitted independently. If
> `cell_type` is supplied, only those labels are processed. OOF folds contain
> condition-stratified cells from the same fitted type; biological sample
> metadata and sample count are not inputs or gates.

# Overview

RegCompassR converts paired single-cell RNA+ATAC evidence into reaction-level
support and directional COMPASS-like scores.

```text
one Pando motif/domain edge dictionary per cell type
→ nested outer-heldout cell projection with inner lambda selection
→ training-only equal-condition TF-RNA × peak-ATAC transforms
→ common-support primary and condition-full exploratory routes
→ supported metabolic genes and complete-GPR cores
→ one ordered biological expansion pass
→ condition × broad cell type SuperCells
→ latent depth-aware RNA and bounded regulatory support
→ separate common/full/RNA/depth/alpha GPR and penalty routes
→ one shared medium-specific GEM
→ directional LP scoring
```

Pando 1.5.0 is the sole GRN estimator. RegCompass consumes its
`ConditionGRNFit v5` contracts. Only outer-heldout projections enter penalties.

# Prepare the human GEM and medium

```{r}
library(RegCompassR)
library(BSgenome.Hsapiens.UCSC.hg38)

gem <- rc_prepare_gem(
  species = "human",
  version = "2.0.0",
  source = "bundled"
)

medium_scenarios <- rc_make_medium_scenarios(
  gem,
  scenario = "high_glucose",
  species = "human"
)
```

When `pfm` is omitted, RegCompass loads Pando's bundled `motifs` object. Human
analyses also use the bundled hg38 phastCons plus SCREEN ccRE union unless
`pando_initiate_args$regions` is supplied.

# Complete workflow

```{r}
result <- rc_run_regcompass_one_shot(
  object = A,
  outdir = "RegCompass_result",
  genome = BSgenome.Hsapiens.UCSC.hg38,
  species = "human",
  gem = gem,
  condition_col = "Group",
  celltype_col = "cell_type",

  pando_args = list(
    min_cells = 100L,
    min_abs_estimate = 0,
    min_model_rsq = 0.1,
    pando_infer_args = list(
      candidate_screen = "motif_domain",
      tf_cor = 0.1,
      peak_cor = 0,
      alpha = 0.5,
      condition_mix = 0.5,
      condition_weight = "equal",
      reference_condition = "Control",
      nlambda = 50L,
      comparison_conditions = c("Control", "Drug"),
      outer_nfolds = 5L,
      inner_nfolds = 5L,
      lambda_selection = "lambda.1se",
      scale = TRUE
    )
  ),

  fragment_files = FALSE,
  metacell_args = list(
    rna_reduction = "pca",
    rna_dims = 1:30,
    atac_reduction = "lsi",
    atac_dims = 2:30,
    gamma = 30L,
    depth_balance = TRUE,
    seed = 12345L,
    min_cells_per_stratum = 500L,
    min_metacell_size = 10L,
    min_metacells_per_stratum = 2L,
    overwrite = FALSE
  ),

  layer1_args = list(
    comparison_support = "auto",
    regulatory_alpha = 0.5,
    gpr_and_method = "min"
  ),
  medium_scenarios = medium_scenarios,
  model_mode = "meta_module_gem",
  layer2_args = list(
    target_direction = "both",
    solver = "highs",
    model_params = list(
      completion_time_limit = 600,
      fastcore_epsilon = 1e-4,
      max_support_reactions = 2000,
      strict = TRUE
    )
  ),
  upstream_workers = 6L,
  layer2_workers = 30L
)
```

The one-shot runner controls Pando execution through `upstream_workers` and a
stage-scoped backend. `pando_infer_args$parallel` and `BPPARAM` are not part of
the nested inference example.

# Pando shared-design contract

For edge `e` and unit `u`, Pando fits the interaction predictor:

```text
x[e, u] = RNA[TF(e), u] × ATAC[peak(e), u]
```

For every outer fold, the interaction center is the mean of condition means
and its scale is the square root of the mean condition-specific population
variance. The transform and lambda are learned only inside that outer training
set. Each cell receives exactly one held-out projection. A separate full-data
refit is used only to interpret the final sub-GRN.

The interaction-safe candidate default is `motif_domain`. The optional
`pooled_within_condition` mode applies response-dependent marginal-correlation
screening and is a sensitivity mode; its projection is ineligible for penalty
construction.

# Interpretation contrasts and primary support

Pando retains full-data edge contrasts for sub-GRN interpretation:

```text
contrast[e, c] = beta[e, c] - beta[e, reference]
comparison_mask[e, c] =
  eligibility_mask[e, c] && eligibility_mask[e, reference]
```

These tables do not drive the primary penalty. The penalty route instead uses
the outer-fold pairwise-common support for two conditions or global-common
support for an omnibus analysis. `condition_estimable` and `strict` are
diagnostic-only. An estimable inactive coefficient remains zero; an
unavailable coefficient remains `NA`.

```{r}
all_effects <- result$grn$tf_peak_gene_condition_effect_all
all_effects[, c(
  "edge_id", "Group", "cell_type",
  "condition_estimate", "reference_estimate", "condition_effect",
  "comparable_to_reference"
)]

table(all_effects$comparable_to_reference, useNA = "ifany")
```

# Pando argument routing

```text
pando_initiate_args → Pando::initiate_grn()
pando_motif_args    → Pando::find_motifs()
pando_infer_args    → Pando::infer_condition_grn()
```

RegCompass owns the object, assays, motif object, genome, metadata columns,
complete GEM target set, network name, minimum condition size, error policy,
and `BPPARAM`. Nested attempts to override these fields stop before Pando is
called. Stage 1 also rejects Pando aggregation columns because Stage 2 owns
metacell construction.

# Restartable Stage 1

```{r}
library(BiocParallel)

upstream_bp <- if (.Platform$OS.type == "windows") {
  SnowParam(workers = 6L, type = "SOCK")
} else {
  MulticoreParam(workers = 6L)
}

step1 <- rc_regcompass_step_grn(
  object = A,
  gem = gem,
  outdir = "RegCompass_steps/01_grn",
  genome = BSgenome.Hsapiens.UCSC.hg38,
  species = "human",
  condition_col = "Group",
  celltype_col = "cell_type",
  pando_args = list(
    min_cells = 100L,
    min_model_rsq = 0.1,
    pando_infer_args = list(
      candidate_screen = "motif_domain",
      reference_condition = "Control"
    )
  ),
  parallel = TRUE,
  BPPARAM = upstream_bp
)

step1$params$pando_parallel
step1$grn_result$condition_grn_fits
step1$grn_result$normalization_policy[c(
  "pando_motifs",
  "pando_regions",
  "pando_candidate_screen",
  "comparison_support",
  "pando_peak_cor",
  "min_model_rsq"
)]
```

With `parallel = TRUE` and `BPPARAM = NULL`, RegCompass enables Pando's native
map backend. With `parallel = FALSE`, Stage 1 is serial. `BPPARAM = TRUE` is
invalid.

# Stages 2–6

```{r}
step2 <- rc_regcompass_step_metacells(
  object = A,
  outdir = "RegCompass_steps/02_metacells",
  condition_col = "Group",
  celltype_col = "cell_type",
  fragment_files = FALSE,
  metacell_args = list(
    rna_reduction = "pca",
    rna_dims = 1:30,
    atac_reduction = "lsi",
    atac_dims = 2:30,
    gamma = 30L,
    depth_balance = TRUE,
    seed = 12345L,
    min_cells_per_stratum = 500L,
    min_metacell_size = 10L,
    min_metacells_per_stratum = 2L
  )
)

step3 <- rc_regcompass_step_meta_modules(
  grn = step1,
  metacells = step2,
  gem = gem,
  outdir = "RegCompass_steps/03_meta_modules"
)

step4 <- rc_regcompass_step_layer1(
  grn = step1,
  metacells = step2,
  meta_modules = step3,
  gem = gem,
  outdir = "RegCompass_steps/04_layer1",
  regulatory_alpha = 0.5,
  gpr_and_method = "min",
  parallel = TRUE,
  BPPARAM = upstream_bp
)

step5 <- rc_regcompass_step_layer2(
  layer1 = step4,
  meta_modules = step3,
  gem = gem,
  medium_scenarios = medium_scenarios,
  outdir = "RegCompass_steps/05_layer2",
  model_mode = "meta_module_gem",
  layer2_args = list(
    target_direction = "both",
    solver = "highs",
    model_params = list(
      completion_time_limit = 600,
      fastcore_epsilon = 1e-4,
      max_support_reactions = 2000,
      strict = TRUE
    )
  )
)

result <- rc_regcompass_step_results(
  grn = step1,
  metacells = step2,
  meta_modules = step3,
  layer1 = step4,
  layer2 = step5,
  gem = gem,
  outdir = "RegCompass_steps/06_results",
  species = "human"
)
```

Stage 3 performs one complete-GPR/subsystem/cross-reference expansion pass and
does not run FASTCORE. Stage 4 aggregates Pando outer-heldout cell projections
through exact SuperCell membership and estimates latent RNA with a
library-offset Gamma-Poisson model. Stage 5 builds each shared structural model
once, then completely reruns GPR, penalty and LP scoring for common OOF,
condition-full OOF, RNA-only, depth, and alpha-sensitivity routes.

The primary modifier is

\[
R_{gm}=q_g\tanh(G^{OOF}_{gm}/s_g),
\qquad
q_g=\sqrt{\operatorname{clamp}(R^2_{OOF,g},0,1)},
\]

where \(s_g\) is estimated jointly across conditions within a cell type.
`regulatory_alpha = 0.5` bounds the regulatory odds multiplier at
\(2^{0.5}\); Stage 4 also records the grid `0, 0.25, 0.5, 1`.

# Audited reaction comparison output

```{r}
result$reaction_comparison_by_metacell[, c(
  "reaction_id", "direction", "medium", "cell_type", "condition",
  "metacell_id", "penalty_rna_only", "penalty_common_oof",
  "penalty_condition_full_oof", "penalty_unique_increment",
  "penalty_per_target_flux", "vmax", "projection_oof_available",
  "common_support_fraction", "condition_full_support_fraction",
  "depth_sensitivity_flag", "zero_support_sensitive",
  "link_saturation_sensitive", "alpha", "inference_class",
  "comparability_class"
)]
```

Common, condition-full, RNA-only, depth-matched and alpha-grid penalties are
produced by complete GPR, penalty and LP reruns on the same GEM, reaction order,
bounds, medium, target direction and \(v_{max}\). The formal quantitative basis
is `penalty / (omega * vmax)`. Within-target ECDF scores are composition
dependent and display-only.

With those structural inputs fixed, the directional LP value is
coordinatewise nondecreasing in its penalty vector, positively 1-homogeneous,
and concave. This establishes the stated mathematical comparison properties;
it does not identify whether an upstream penalty difference is biological or
technical. Depth, zero-pattern, link-saturation and alpha sensitivity results
must be interpreted alongside the primary common-OOF result.

Metacells are the statistical units for within-dataset Wilcoxon and
Kruskal--Wallis comparisons, with multiplicity-adjusted P values and effect
sizes retained. They are not sample/donor-level biological replicates, so these
tests do not constitute biological-replicate-level treatment inference.

# Mouse input

Pando's bundled regulatory-region objects are hg38. Mouse runs must supply a
build-matched `GRanges` object:

```{r}
library(BSgenome.Mmusculus.UCSC.mm10)

mouse_regions <- readRDS("mm10_regulatory_regions.rds")

mouse_step1 <- rc_regcompass_step_grn(
  object = A_mouse,
  gem = mouse_gem,
  outdir = "RegCompass_mouse/01_grn",
  genome = BSgenome.Mmusculus.UCSC.mm10,
  species = "mouse",
  condition_col = "Group",
  celltype_col = "cell_type",
  pando_args = list(
    pando_initiate_args = list(regions = mouse_regions),
    pando_infer_args = list(
      candidate_screen = "motif_domain",
      reference_condition = "Control"
    )
  )
)
```

RegCompass stops instead of silently applying hg38 regions to mouse ATAC peaks.


# Single-sample conditions and SuperCell strata

SuperCell hard strata are exactly condition × broad cell type. Biological sample count is not an analysis requirement: single- and multi-sample conditions use the same paired RNA+ATAC path and the same condition-stratified cell OOF within each fitted cell type.
