---
title: "RegCompassR 1.7.0 condition-pooled workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RegCompassR 1.7.0 condition-pooled workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

RegCompassR 1.7.0 uses ATAC and RNA from cells pooled across biological samples
within each condition and cell type. SuperCell2 builds condition-level
metacells, Pando infers condition-specific TF–peak–gene regulation, and one
shared union-GEM is used for directional COMPASS-like scoring.

## Prepare the model

```{r}
library(RegCompassR)

human_gem <- rc_prepare_human2_gem(version = "2.0.0")
mouse_gem <- rc_prepare_mouse_gem(version = "1.8.0")
gem <- rc_prepare_gem(species = "human", version = "2.0.0")

medium <- rc_make_medium_scenarios(
  human_gem,
  scenario = "normal_human_plasma"
)
```

## Run the explicit workflow

```{r}
library(Pando)
library(BSgenome.Hsapiens.UCSC.hg38)
data(motifs, package = "Pando")
data(SCREEN.ccRE.UCSC.hg38, package = "Pando")

result <- rc_run_regcompass(
  object = object,
  gem = human_gem,
  outdir = "RegCompass_result",
  pfm = motifs,
  genome = BSgenome.Hsapiens.UCSC.hg38,
  fragment_files = FALSE,
  species = "human",
  sample_col = "sample_id",
  condition_col = "condition",
  celltype_col = "cell_type",
  medium_scenarios = medium,
  inference_unit = "metacell",
  metacell_args = list(
    gamma = 150,
    min_cells_per_stratum = 500,
    min_metacell_size = 10
  ),
  pando_args = list(
    min_metacells = 10,
    pando_initiate_args = list(
      regions = SCREEN.ccRE.UCSC.hg38
    ),
    pando_infer_args = list(
      method = "glm",
      tf_cor = 0.1,
      peak_cor = 0.05,
      adjust_method = "fdr"
    )
  ),
  layer1_args = list(
    regulatory_alpha = 1,
    tau = 0.20,
    local_fastcore = TRUE
  ),
  layer2_args = list(
    target_direction = "both",
    solver = "highs"
  )
)
```

The main outputs are the integrated gene and reaction evidence in
`result$layer1`, the union-GEM membership in `result$grn_meta_modules`, raw
minimum penalties in `result$microcompass$penalty`, and descriptive condition
comparisons in `result$condition_contrast`.

## Equivalent one-shot entry point

```{r}
result <- rc_run_regcompass_one_shot(
  object = object,
  outdir = "RegCompass_result",
  pfm = motifs,
  genome = BSgenome.Hsapiens.UCSC.hg38,
  fragment_files = FALSE,
  species = "human",
  gem_version = "2.0.0",
  medium_scenario = "normal_human_plasma",
  sample_col = "sample_id",
  condition_col = "condition",
  celltype_col = "cell_type",
  inference_unit = "metacell",
  pando_args = list(
    min_metacells = 10,
    pando_initiate_args = list(
      regions = SCREEN.ccRE.UCSC.hg38
    )
  )
)
```

For mouse data, use `species = "mouse"`, a mouse genome object, and Mouse-GEM.
