---
title: "RegCompassR workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RegCompassR workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

# Overview

RegCompassR converts paired single-cell RNA+ATAC data into regulatory reaction
support and directional metabolic scores. It automatically selects one of two
Pando routes:

```text
>=2 condition levels -> canonical condition_grn with outer-heldout projection
0 or 1 condition level -> standard_pando using original Pando::infer_grn()
```

The standard route does not calculate condition coefficients. It projects the
standard TF–peak–gene coefficients as cell-level TF RNA × peak ATAC terms and
continues through the same metacell, GPR, penalty, and metabolic-model stages.

SuperCell receives separate graph and purity inputs:

```text
cell.graph.group = broad cell type
cell.split.condition = condition, or NULL when omitted
```

Each broad cell type is built as an independent graph. All conditions within
that cell type are standardized and embedded jointly, and condition is applied
only after graph clustering to keep final metacells condition-pure. No sample or
temporary combined condition-by-cell-type field is used.

Mathematical definitions are centralized in
[Mathematical model](../docs/mathematical-model.md) and the
[metacell graph contract](../docs/metacell-graph-contract.md).

# Prepare GEM and medium

```{r}
library(RegCompassR)
library(BSgenome.Hsapiens.UCSC.hg38)

gem <- rc_prepare_gem(
  species = "human",
  version = "2.0.0",
  source = "bundled"
)

medium_scenarios <- rc_make_medium_scenarios(
  gem,
  scenario = "physiologic",
  species = "human"
)
```

# Multiple-condition workflow

```{r}
result <- rc_run_regcompass(
  object = A,
  gem = gem,
  outdir = "RegCompass_result",
  genome = BSgenome.Hsapiens.UCSC.hg38,
  species = "human",
  condition_col = "Group",
  celltype_col = "cell_type",
  pando_args = list(
    min_cells = 100L,
    min_model_rsq = 0.1,
    pando_infer_args = list(
      candidate_screen = "motif_domain",
      condition_mix = 0.5,
      condition_weight = "equal",
      outer_nfolds = 5L,
      inner_nfolds = 5L,
      scale = TRUE
    )
  ),
  metacell_args = list(
    rna_reduction = "pca",
    rna_dims = 1:30,
    atac_reduction = "lsi",
    atac_dims = 2:30,
    gamma = 30L,
    seed = 12345L
  ),
  layer1_args = list(
    comparison_support = "auto",
    regulatory_alpha = 1,
    gpr_and_method = "min"
  ),
  medium_scenarios = medium_scenarios,
  model_mode = "meta_module_gem"
)
```

This route returns `result$analysis_mode == "condition_grn"` and uses the
canonical unversioned `pando_condition_grn_fit` contract. Stage 2 records:

```r
result$metacells$input_design$graph_scope
result$metacells$input_design$condition_scope
```

with values `"one_independent_graph_per_cell_type"` and
`"all_conditions_joint_within_cell_type_graph"`.

# Standard Pando workflow

Omit the condition column or provide one level:

```{r}
standard_result <- rc_run_regcompass(
  object = A,
  gem = gem,
  outdir = "RegCompass_standard",
  genome = BSgenome.Hsapiens.UCSC.hg38,
  species = "human",
  condition_col = NULL,
  celltype_col = "cell_type",
  metacell_args = list(
    rna_reduction = "pca",
    atac_reduction = "lsi",
    gamma = 30L
  )
)

standard_result$analysis_mode
standard_result$condition_coefficients_calculated
standard_result$reaction_ranking
```

The expected values are `"standard_pando"` and `FALSE`. The condition contrast
is empty because no second condition is manufactured.

# Stepwise workflow

```{r}
step1 <- rc_regcompass_step_grn(
  object = A,
  gem = gem,
  outdir = "steps/01_grn",
  genome = BSgenome.Hsapiens.UCSC.hg38,
  condition_col = "Group",
  celltype_col = "cell_type"
)

step2 <- rc_regcompass_step_metacells(
  object = A,
  outdir = "steps/02_metacells",
  condition_col = "Group",
  celltype_col = "cell_type",
  metacell_args = list(
    rna_reduction = "pca",
    atac_reduction = "lsi",
    gamma = 30L
  )
)

step3 <- rc_regcompass_step_meta_modules(
  grn = step1,
  metacells = step2,
  gem = gem,
  outdir = "steps/03_meta_modules"
)

step4 <- rc_regcompass_step_layer1(
  grn = step1,
  metacells = step2,
  meta_modules = step3,
  gem = gem,
  outdir = "steps/04_layer1",
  comparison_support = "auto",
  regulatory_alpha = 1,
  gpr_and_method = "min"
)

step5 <- rc_regcompass_step_layer2(
  layer1 = step4,
  meta_modules = step3,
  gem = gem,
  medium_scenarios = medium_scenarios,
  outdir = "steps/05_layer2"
)

result <- rc_regcompass_step_results(
  grn = step1,
  metacells = step2,
  meta_modules = step3,
  layer1 = step4,
  layer2 = step5,
  gem = gem,
  outdir = "steps/06_results"
)
```

Stage 1 and Stage 2 independently resolve the analysis mode and must agree.

# Interpretation

- condition mode: primary regulatory evidence is outer-heldout common-support
  projection;
- standard mode: regulatory evidence comes from the original Pando full fit;
- both modes calculate TF×ATAC before metacell aggregation;
- each cell type has an independent graph and all its conditions share that
  graph before condition-pure membership splitting;
- non-finite target modifiers use neutral `R=0`, exactly matching RNA-only
  support for that entry;
- `regulatory_alpha = 1` is fixed;
- missing final reaction expression is converted to `E=0` and receives the
  maximum expression-linked penalty;
- metacells are within-dataset statistical units, not biological replicates.

# Further documentation

- [Workflow](../docs/workflow.md)
- [Run modes](../docs/run-modes-and-stepwise-workflow.md)
- [Stage contracts](../docs/stage-interface-contracts.md)
- [Metacell graph contract](../docs/metacell-graph-contract.md)
- [Public API](../docs/functions.md)
