---
title: "RegCompassR workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RegCompassR workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

# Overview

RegCompassR converts paired single-cell RNA+ATAC data into regulatory reaction
support and directional metabolic scores.

```text
>=2 condition levels → common-dictionary condition GRNs
0 or 1 condition level → original per-cell-type Pando GRN
```

For multiple conditions, each broad cell type uses one two-stage Pando design:

```text
global candidate discovery
+ condition-specific candidate discovery
→ exact TF-peak-target union
→ frozen target-specific dictionary
→ one unscaled Gaussian identity GLM per condition
→ BH-adjusted condition effects
```

The first-stage candidate fits are not final effects. No pooled coefficient is
used to rescale a condition coefficient. Conditions may have different
significant edges and opposite directions, while all coefficients remain on the
same predictor definition and numerical scale.

Stage 2 calls `SuperCell::SCimplify_by_graph_group()` with one independent
multimodal WNN graph per broad cell type. All conditions of that type jointly
determine the graph; condition splits parent membership only after clustering.
Final metacells are therefore condition-pure without condition-specific graphs.

# Run

```{r}
library(RegCompassR)
library(BSgenome.Hsapiens.UCSC.hg38)

gem <- rc_prepare_gem(
  species = "human",
  version = "2.0.0",
  source = "bundled"
)

medium_scenarios <- rc_make_medium_scenarios(
  gem,
  scenario = "normal_human_plasma",
  species = "human"
)

result <- rc_run_regcompass(
  object = A,
  gem = gem,
  outdir = "RegCompass_result",
  genome = BSgenome.Hsapiens.UCSC.hg38,
  species = "human",
  condition_col = "Group",
  celltype_col = "cell_type",
  pando_args = list(
    min_cells = 300L,
    min_abs_estimate = 0,
    min_model_rsq = 0.1,
    pando_infer_args = list(
      tf_cor = 0.1,
      peak_cor = 0,
      adjust_method = "BH",
      padj_threshold = 0.05,
      rank_action = "mark",
      min_residual_df = 1L
    )
  ),
  metacell_args = list(
    rna_reduction = "pca",
    rna_dims = 1:30,
    atac_reduction = "lsi",
    atac_dims = 2:30,
    gamma = 30L,
    k.knn = 30L,
    seed = 12345L
  ),
  layer1_args = list(
    projection_component = "condition",
    regulatory_alpha = 1,
    gpr_and_method = "min"
  ),
  medium_scenarios = medium_scenarios,
  model_mode = "meta_module_gem"
)
```

Stage 1 fixes `min_cells` at 300. The following former condition-GRN parameters
are rejected:

```text
candidate_screen
condition_mix
condition_weight
alpha
nlambda / lambda / lambda_min_ratio
outer_nfolds / inner_nfolds
lambda_selection
scale
engine_control
comparison_conditions
```

# Condition effect and penalty

For an estimable edge, `estimate` is the complete fixed-dictionary condition
coefficient. Pando calculates a network-wide BH-adjusted P value and defines:

```text
penalty_effect = estimate when padj < 0.05
penalty_effect = 0 otherwise
```

RegCompass then applies `min_abs_estimate` and `min_model_rsq`. The complete
coefficient table is retained even when an edge does not enter penalty. A
zero-variance or aliased coefficient remains `NA` and is not interpreted as a
fitted zero.

The GLM P value is conditional on the frozen candidate dictionary and does not
include selective-inference correction for candidate discovery.

# Cell-first projection

Pando reconstructs each paired-cell predictor `TF RNA × peak ATAC`, applies the
condition `penalty_effect`, and sums edge contributions by target. RegCompass
then aggregates these target scores by exact SuperCell membership. It does not
recompute TF×ATAC from metacell averages or refit after aggregation.

```{r}
result$grn$tf_peak_gene_condition_effect_all
result$grn$tf_peak_gene_condition_effect
result$layer1$gene_regulatory_modifier
result$microcompass$penalty
result$reaction_ranking
result$condition_contrast
```

Several historical fields retain `_oof`, `common`, or `condition_unique` in their
names for backward compatibility. In the current method, the primary and common
fields are aliases of the BH-filtered fixed-dictionary full-fit projection, and
the condition-unique decomposition is zero. The names do not imply OOF fitting
or a common-support model.

# No condition or one condition

With no usable condition column or one observed condition level, RegCompass
runs original Pando independently within each retained broad cell type. It does
not create condition coefficients or a synthetic condition fit contract.

# Optional targeted reaction remapping

After a completed `model_mode = "meta_module_gem"` run, directly database-linked
non-core reactions can be scored without rebuilding the model:

```{r}
targeted <- rc_regcompass_step_target_union(
  layer1 = result$layer1,
  meta_modules = result$meta_modules,
  layer2 = result$microcompass,
  gem = gem,
  outdir = "RegCompass_targeted",
  core_reaction_ids = c("MAR04381", "MAR04379"),
  layer2_args = list(target_direction = "both", solver = "highs")
)
```

The second pass reuses the same Stage 5 union GEM, global FASTCORE completion,
reaction order, direction, bounds and `vmax`.

# Documentation

- [Tutorial 1](../docs/tutorial-01-quick-start.md)
- [Tutorial 2](../docs/tutorial-02-stepwise-audit.md)
- [Condition-GRN contract](../docs/condition-comparable-grn.md)
- [Tutorial 4: targeted reaction remapping](../docs/tutorial-04-targeted-reaction-remapping.md)
- [Tutorial 5: condition comparison](../docs/tutorial-05-condition-differential-analysis.md)
- [Medium scenarios and evidence](../docs/medium-presets.md)
- [Stage contracts](../docs/stage-interface-contracts.md)
- [Public API](../docs/functions.md)
