---
title: "RegCompassR 1.7.0 condition-pooled workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RegCompassR 1.7.0 condition-pooled workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

RegCompassR 1.7.0 pools cells across biological samples within each condition and
cell type. SuperCell2 builds condition-level metacells while retaining original
sample composition. Pando fits one condition-specific TF–peak–gene model per cell
type. Directional COMPASS-like scoring can use either the shared union meta-module
GEM or the shared full GEM.

One or more samples per condition and one or more conditions are accepted.
Condition-pooled metacells are descriptive units, not independent biological
replicates. A single-condition analysis returns a metabolic reaction ranking;
multiple conditions additionally return pairwise descriptive priority contrasts.

Pando coefficients are learned from RNA+ATAC, but the per-metacell regulatory
state uses coefficient-weighted peak accessibility only. Metacell TF RNA is not
multiplied into that state.

Biological meta-module membership is annotation-defined: complete-GPR core
reactions are expanded through their subsystems and shared KEGG, Reactome, or
master-Rhea reaction identifiers. No metabolite-neighbour or stoichiometric
one-hop expansion is performed. Local FASTCORE is the sole stage that can add
non-annotated reactions required for flux feasibility, and those support
reactions remain separate from biological membership.

## Prepare the model

```{r}
library(RegCompassR)

human_gem <- rc_prepare_human2_gem(version = "2.0.0")
mouse_gem <- rc_prepare_mouse_gem(version = "1.8.0")
gem <- rc_prepare_gem(species = "human", version = "2.0.0")

medium <- rc_make_medium_scenarios(
  human_gem,
  scenario = "normal_human_plasma"
)
```

## Run the explicit workflow

```{r}
library(Pando)
library(BSgenome.Hsapiens.UCSC.hg38)
data(motifs, package = "Pando")
data(SCREEN.ccRE.UCSC.hg38, package = "Pando")

result <- rc_run_regcompass(
  object = object,
  gem = human_gem,
  outdir = "RegCompass_result",
  pfm = motifs,
  genome = BSgenome.Hsapiens.UCSC.hg38,
  fragment_files = FALSE,
  species = "human",
  sample_col = "sample_id",
  condition_col = "condition",
  celltype_col = "cell_type",
  model_mode = "meta_module_gem",
  medium_scenarios = medium,
  metacell_args = list(
    gamma = 150,
    min_cells_per_stratum = 500,
    min_metacell_size = 10
  ),
  pando_args = list(
    min_metacells = 10,
    pando_initiate_args = list(
      regions = SCREEN.ccRE.UCSC.hg38
    ),
    pando_infer_args = list(
      method = "glm",
      tf_cor = 0.1,
      peak_cor = 0.05,
      adjust_method = "fdr"
    )
  ),
  layer1_args = list(
    regulatory_alpha = 1,
    tau = 0.20,
    local_fastcore = TRUE
  ),
  layer2_args = list(
    target_direction = "both",
    solver = "highs"
  )
)
```

Use `model_mode = "full_gem"` to retain the complete validated GEM during LP
scoring. The same condition-pooled Layer 1 evidence, medium, core-derived targets,
and output ranking contract are used in both modes.

The main outputs are:

- sample membership in `result$metacells$sample_composition`;
- RNA, ATAC-derived regulatory state and `reaction_expression` in `result$layer1`;
- biological membership, separate FASTCORE support and shared union membership in `result$grn_meta_modules`;
- raw minimum penalties in `result$microcompass$penalty`;
- reaction priorities in `result$reaction_ranking`;
- pairwise descriptive comparisons in `result$condition_contrast` when multiple conditions are present.

With one condition, `result$analysis_mode` is
`"single_condition_reaction_ranking"` and `result$condition_contrast` is empty.

The result provenance records
`result$params$meta_module_expansion = "core_subsystem_plus_kegg_reactome_master_rhea_only"`
and `result$params$feasibility_completion = "local_fastcore_only"`.

## Equivalent one-shot entry point

```{r}
result <- rc_run_regcompass_one_shot(
  object = object,
  outdir = "RegCompass_result",
  pfm = motifs,
  genome = BSgenome.Hsapiens.UCSC.hg38,
  fragment_files = FALSE,
  species = "human",
  gem_version = "2.0.0",
  medium_scenario = "normal_human_plasma",
  sample_col = "sample_id",
  condition_col = "condition",
  celltype_col = "cell_type",
  model_mode = "meta_module_gem",
  pando_args = list(
    min_metacells = 10,
    pando_initiate_args = list(
      regions = SCREEN.ccRE.UCSC.hg38
    )
  )
)
```

The one-shot entry point delegates to the same condition-pooled evidence,
annotation-only meta-module architecture, and selectable structural model modes.

For mouse data, use `species = "mouse"`, a mouse genome object, and Mouse-GEM.
