---
title: "RegCompassR 1.8.9 shared-background condition-sub-GRN workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RegCompassR 1.8.9 shared-background condition-sub-GRN workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

# Overview

RegCompassR 1.8.9 separates a shared structural regulatory hypothesis from
condition-specific sparse regulatory coefficients.

```text
one validated GREAT-domain Pando TF–peak–target universe per cell type
→ condition-aware TF×peak/target observability filter
→ one shared model edge universe across conditions
→ direct condition-specific theta elastic net
→ derived cross-condition backbone and zero-sum deviations
→ leakage-resistant condition-stratified cross-validation
→ full-size condition-stratified bootstrap reproducibility
→ condition-specific metabolic target genes
→ complete-GPR condition cores
→ one ordered annotation expansion pass
→ one merged reaction catalogue
→ one medium-specific union GEM and one global FASTCORE completion
→ identical stoichiometric structure across conditions
→ condition/metacell-specific penalties and directional LP scores
→ compact final result
```

For edge \(e\), RegCompass directly estimates \(\theta_{e,c}\). The exported
backbone and deviation are derived:

\[
\beta_e=\frac1C\sum_c\theta_{e,c},
\qquad
\delta_{e,c}=\theta_{e,c}-\beta_e,
\qquad
\sum_c\delta_{e,c}=0.
\]

Every condition uses the same Pando structural dictionary and the same
observability-filtered model dictionary. Each condition contributes the same
total regression loss weight. The canonical public workflow uses condition and
cell type metadata only; it does not accept a biological-sample column.

# Prepare the GEM and medium

```{r}
library(RegCompassR)
library(BSgenome.Hsapiens.UCSC.hg38)

gem <- rc_prepare_gem(
  species = "human",
  version = "2.0.0",
  source = "bundled"
)

medium_scenarios <- rc_make_medium_scenarios(
  gem,
  scenario = "high_glucose",
  species = "human"
)
```

When `pfm` is omitted, RegCompass loads `data("motifs", package = "Pando")`.
Human analyses use phastCons plus SCREEN ccRE regions; mouse analyses use the
available phastCons resource unless overridden.

# Complete workflow

```{r}
result <- rc_run_regcompass_one_shot(
  object = A,
  outdir = "RegCompass_result",
  genome = BSgenome.Hsapiens.UCSC.hg38,
  species = "human",
  gem = gem,
  condition_col = "Group",
  celltype_col = "cell_type",

  grn_mode = "multitask_shared_backbone",
  pando_args = list(
    min_cells = 100,
    pando_design_args = list(
      peak_to_gene_method = "GREAT",
      upstream = 100000,
      downstream = 0,
      extend = 1000000,
      only_tss = FALSE,
      min_tf_detection = 0,
      min_peak_detection = 0,
      min_target_detection = 0,
      max_edges_per_target = Inf
    )
  ),
  multitask_args = list(
    alpha = 0.5,
    global_penalty_factor = 1,
    deviation_penalty_factor = 1,
    lambda_rule = "lambda.1se",
    nfolds = 5,
    n_bootstrap = 100,
    min_selection_frequency = 0.7,
    min_sign_stability = 0.8,
    min_bootstrap_success_fraction = 0.8,
    min_cv_rsq = 0,
    candidate_screen_threshold = 0,
    max_edges_per_target = Inf,
    min_detected_cells_per_condition = 10,
    min_detection_fraction_per_condition = 0.01,
    seed = 12345L
  ),

  fragment_files = FALSE,
  metacell_args = list(
    rna_reduction = "pca",
    rna_dims = 1:30,
    atac_reduction = "lsi",
    atac_dims = 2:30,
    gamma = 30,
    seed = 12345L,
    min_cells_per_stratum = 500,
    min_metacell_size = 10,
    min_metacells_per_stratum = 2L,
    overwrite = FALSE
  ),

  layer1_args = list(
    regulatory_alpha = 1,
    gpr_and_method = "min"
  ),

  medium_scenarios = medium_scenarios,
  model_mode = "meta_module_gem",
  layer2_args = list(
    target_direction = "both",
    solver = "highs",
    model_params = list(
      completion_time_limit = 600,
      fastcore_epsilon = 1e-4,
      max_support_reactions = 2000,
      strict = TRUE
    )
  ),
  upstream_workers = 6,
  layer2_workers = 30
)
```

# GREAT structural candidate universe

`Pando::prepare_grn_design()` constructs structural candidates before fitting.
The canonical `peak_to_gene_method = "GREAT"` creates basal-plus-extension
regulatory domains and permits distal hypotheses up to `extend = 1000000`.
Candidate admission does not use fitted target-expression correlation,
coefficients, P values, or condition effects.

An edge requires a target-linked regulatory region, a matched TF motif, an exact
measured ATAC feature, and measured TF/target RNA features. Predictors are
deduplicated by `(TF, ATAC feature, target)`, and the version-2 design
fingerprint is validated before use. `Signac` remains an explicit
narrower-domain sensitivity option.

The predictor is

\[
x_{e,u}=T_{t,u}A_{p,u}.
\]

Pando pooled detection thresholds remain zero so condition-restricted
candidates are not removed prematurely. RegCompass defines

\[
m_c=\min\left(n_c,\max\left(10,\left\lceil0.01n_c\right\rceil\right)\right)
\]

and retains an edge for modelling only when, in at least one condition,

\[
\sum_{u\in c}I(T_{t,u}>0\land A_{p,u}>0)\ge m_c
\]

and

\[
\sum_{u\in c}I(Y_{g,u}>0)\ge m_c.
\]

TF and peak must therefore be observed together in the same cells. This
detection filter does not use target correlation or fitted coefficients. The
complete structural universe and filtered model universe receive separate
fingerprints.

# Direct condition-theta model and predictive validation

Target and predictor values are centred within condition. Predictor scales are
shared across conditions, while observation weights give every condition equal
total loss. The model is

\[
y_u^\circ=\sum_e\widetilde x_{e,u}\theta_{e,c(u)}+\varepsilon_u.
\]

The elastic-net objective penalises \(\theta_{e,c}\) directly:

\[
\min_\Theta
\frac12\sum_u w_u
\left(y_u^\circ-\sum_e\widetilde x_{e,u}\theta_{e,c(u)}\right)^2
+\lambda\alpha p_\theta\sum_{e,c}|\theta_{e,c}|
+\frac{\lambda(1-\alpha)p_\theta}{2}\sum_{e,c}\theta_{e,c}^2.
\]

Thus a fitted edge may satisfy

\[
\theta_{e,A}\ne0,\qquad\theta_{e,B}=0.
\]

`global_penalty_factor` and `deviation_penalty_factor` are retained as
compatibility aliases for the same \(p_\theta\) and must be equal.

For every cross-validation fold, condition centres and edge scales are estimated
from training cells and applied to validation cells. Lambda uses the
one-standard-error rule. A target model can produce active edges only when

\[
R^2_{CV}>0.
\]

A positive `min_cv_rsq` adds a stronger predictive floor.

# Bootstrap reproducibility

For each bootstrap replicate, condition \(c\) is resampled with replacement at
its original size \(n_c\). The target and predictors are re-centred within
condition, scaled by the shared full-data scale, and fitted at the selected
lambda.

For successful bootstrap set \(\mathcal B_s\),

\[
\Pi_{e,c}=\frac1{|\mathcal B_s|}
\sum_{b\in\mathcal B_s}I(|\theta^{(b)}_{e,c}|>\varepsilon),
\]

\[
\rho_{e,c}=\left|
\frac{\sum_bI^{(b)}_{e,c}\operatorname{sign}(\theta^{(b)}_{e,c})}
{\sum_bI^{(b)}_{e,c}}
\right|,
\]

\[
\widetilde\theta_{e,c}
=\widehat\theta_{e,c}\Pi_{e,c}\rho_{e,c}.
\]

The canonical default is 100 bootstrap replicates. RegCompass reports
selection-frequency Monte Carlo standard error and Wilson 95% intervals.
`min_sign_stability = 0.8` requires at least 90% majority-sign agreement among
selected fits because \(\rho=|2q-1|\).

Regularized coefficients do not receive classical adjusted P values; `padj`
remains `NA`. The bootstrap quantifies cell-resampling reproducibility, not
biological-replicate significance.

# Condition genes and complete-GPR cores

A gene enters the condition target set when at least one active edge targets it.
Positive and negative directions both establish membership.

For reaction \(r\) with GPR branches \(B_{r,k}\):

\[
Core_{r,c}=1
\iff
\exists k:B_{r,k}\subseteq G_c.
\]

Stage 3 then performs exactly one ordered expansion:

```text
complete-GPR cores
→ core-reaction subsystems
→ direct KEGG/Reactome equivalents
→ direct master-Rhea equivalents
→ stop
```

# Layer 1 and shared metabolic model

Stage 1 uses RNA and ATAC to fit the GRN. Stage 4 reprojects only metacell ATAC.
TF edges sharing one measured peak are signed-summed before projection. One
target denominator is shared across conditions.

A gene with no active condition edge has `R = 0`, and the log-odds update returns
RNA-only support exactly:

\[
C^{MO}=
\frac{C^{RNA}2^{\alpha R}}
{1-C^{RNA}+C^{RNA}2^{\alpha R}}.
\]

Condition core sets may differ, but Stage 3 merges their reaction catalogues
before Stage 5. For each medium, Stage 5 constructs one union GEM and runs the
only global FASTCORE completion. All conditions and metacells reuse identical
reaction IDs, stoichiometric matrix, and bounds.

`completion_time_limit` limits FASTCORE union-GEM construction only.
Directional scoring LPs have no time-limit parameter.

# Stepwise inspection

```{r}
step1 <- rc_regcompass_step_grn(
  object = A,
  gem = gem,
  outdir = "RegCompass_steps/01_grn",
  genome = BSgenome.Hsapiens.UCSC.hg38,
  species = "human",
  condition_col = "Group",
  celltype_col = "cell_type",
  grn_mode = "multitask_shared_backbone",
  pando_args = list(
    pando_design_args = list(peak_to_gene_method = "GREAT")
  )
)

step2 <- rc_regcompass_step_metacells(
  object = A,
  outdir = "RegCompass_steps/02_metacells",
  condition_col = "Group",
  celltype_col = "cell_type",
  fragment_files = FALSE
)

step3 <- rc_regcompass_step_meta_modules(
  grn = step1,
  metacells = step2,
  gem = gem,
  outdir = "RegCompass_steps/03_meta_modules"
)

step4 <- rc_regcompass_step_layer1(
  metacells = step2,
  meta_modules = step3,
  gem = gem,
  outdir = "RegCompass_steps/04_layer1",
  regulatory_alpha = 1,
  gpr_and_method = "min"
)

step5 <- rc_regcompass_step_layer2(
  layer1 = step4,
  meta_modules = step3,
  gem = gem,
  medium_scenarios = medium_scenarios,
  outdir = "RegCompass_steps/05_layer2",
  model_mode = "meta_module_gem",
  layer2_args = list(target_direction = "both", solver = "highs")
)

result <- rc_regcompass_step_results(
  grn = step1,
  metacells = step2,
  meta_modules = step3,
  layer1 = step4,
  layer2 = step5,
  gem = gem,
  outdir = "RegCompass_steps/06_results",
  species = "human"
)
```

Detailed Stage 1 outputs remain in `step1`; compact analysis tables remain in
`result`. See `docs/grn-parameter-policy.md` for the canonical parameter
contract and `docs/tutorial-03-advanced-restart.md` for GREAT-versus-Signac
structural sensitivity.
