---
title: "RegCompassR 1.8.2 GRN-first workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RegCompassR 1.8.2 GRN-first workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

RegCompassR fits condition-by-cell-type Pando GRNs, builds condition-level SuperCell2 metacells, maps complete-GPR core reactions, constructs a FASTCORE-completed global union GEM, integrates RNA and ATAC evidence, and performs directional minimum-penalty LP scoring.

Version 1.8.2 adds strict stage provenance and an optional second LP pass for non-core reactions directly sharing KEGG, Reactome, or master-Rhea identifiers with selected cores.

## Installation

```{r}
install.packages(c("remotes", "highs", "BiocManager", "ggplot2"))
BiocManager::install("BiocParallel", ask = FALSE, update = FALSE)
remotes::install_github("1667857557/Pando_regcompass", upgrade = "never")
remotes::install_github("1667857557/Regcompass", upgrade = "never")
```

A local Pando archive is also accepted:

```{r}
install.packages("~/Pando_regcompass.tar.gz", repos = NULL, type = "source")
```

## Input and workers

The Seurat object must contain paired RNA and ATAC counts for the same cells, complete condition/cell-type metadata, and an ATAC `ChromatinAssay`. Use `Pando::motifs`; do not pass `motif2tf` as `pfm`.

```{r}
library(RegCompassR)
library(Pando)
library(BiocParallel)
library(BSgenome.Hsapiens.UCSC.hg38)

data(motifs, package = "Pando")
condition_col <- "dataset"
celltype_col <- "epithelial_or_stem"

gem <- rc_prepare_gem(species = "human", version = "2.0.0")
medium_scenarios <- rc_make_medium_scenarios(
  gem = gem,
  scenario = "physiologic",
  species = "human"
)

upstream_bp <- MulticoreParam(workers = 16L, progressbar = TRUE)
layer2_bp <- MulticoreParam(workers = 12L, progressbar = TRUE)
```

Set `OMP_NUM_THREADS=1`, `OPENBLAS_NUM_THREADS=1`, and `MKL_NUM_THREADS=1` before starting a multicore R session.

## Stage 1: GRNs

```{r}
step1 <- rc_regcompass_step_grn(
  object = A,
  gem = gem,
  outdir = "steps/01_grn",
  pfm = motifs,
  genome = BSgenome.Hsapiens.UCSC.hg38,
  condition_col = condition_col,
  celltype_col = celltype_col,
  pando_args = list(
    min_cells = 100,
    pando_infer_args = list(
      method = "glm",
      tf_cor = 0.1,
      peak_cor = 0.01,
      adjust_method = "fdr",
      parallel = FALSE
    )
  ),
  parallel = TRUE,
  BPPARAM = upstream_bp
)

stopifnot(
  all(step1$grn_result$sample_status$status == "ok"),
  nzchar(step1$gem_fingerprint)
)
```

RNA is normalized once. ATAC TF-IDF is calculated within each cell type across conditions. Pando is then fitted for each condition × cell-type group.

## Stage 2: metacells

```{r}
step2 <- rc_regcompass_step_metacells(
  object = A,
  outdir = "steps/02_metacells",
  condition_col = condition_col,
  celltype_col = celltype_col,
  fragment_files = FALSE,
  metacell_args = list(
    gamma = 75,
    min_cells_per_stratum = 500,
    min_metacell_size = 10
  )
)

stopifnot(
  !any(step2$pooled$metacell_meta$dominant_celltype_tied %in% TRUE)
)
```

Condition is the hard stratum. The existing cell-type annotation guides SuperCell2 before aggregation and is audited afterwards from member-cell composition.

## Stage 3: core reactions and meta-modules

```{r}
step3 <- rc_regcompass_step_meta_modules(
  grn = step1,
  metacells = step2,
  gem = gem,
  outdir = "steps/03_meta_modules",
  layer1_args = list(
    local_fastcore = TRUE,
    local_fastcore_args = list(
      solver = "highs",
      strict = TRUE,
      parallel = TRUE,
      workers = 16L,
      backend = "multicore"
    )
  )
)

stopifnot(
  all(step3$group_coverage$coverage_complete),
  nrow(step3$global_modules$global_core_reactions) > 0
)
```

Meta-modules contain complete-GPR cores, same-subsystem reactions, and reactions sharing KEGG, Reactome, or master-Rhea identifiers. FASTCORE adds separately labelled feasibility support.

## Stage 4 and Stage 5

```{r}
step4 <- rc_regcompass_step_layer1(
  metacells = step2,
  meta_modules = step3,
  gem = gem,
  outdir = "steps/04_layer1",
  parallel = TRUE,
  BPPARAM = upstream_bp
)

step5 <- rc_regcompass_step_layer2(
  layer1 = step4,
  meta_modules = step3,
  gem = gem,
  medium_scenarios = medium_scenarios,
  outdir = "steps/05_layer2",
  model_mode = "meta_module_gem",
  layer2_args = list(target_direction = "both", solver = "highs"),
  parallel = TRUE,
  BPPARAM = layer2_bp
)

stopifnot(
  inherits(step4, "regcompass_layer1_step"),
  inherits(step5, "regcompass_layer2_step"),
  identical(step4$gem_fingerprint, step5$gem_fingerprint),
  identical(colnames(step4$reaction_expression), colnames(step5$penalty)),
  any(step5$evaluated)
)
```

The raw minimum penalty is the primary result. Lower values indicate stronger compatibility between the required target flux and integrated evidence. The relative score is a within-target rank, not a probability.

## Direct database-linked non-core scoring

```{r}
expanded <- rc_regcompass_step_target_union(
  layer1 = step4,
  meta_modules = step3,
  layer2 = step5,
  gem = gem,
  outdir = "steps/05b_expanded_targets",
  core_genes = c("GCLC", "GCLM", "GSS", "GSR", "G6PD", "PGD"),
  gene_match = "complete_gpr",
  parallel = TRUE,
  BPPARAM = layer2_bp
)

stopifnot(
  all(expanded$expanded_scoring_targets$score_target),
  all(expanded$expanded_reaction_catalog$expansion_type %in% c(
    "shared_kegg_reaction",
    "shared_reactome_reaction",
    "shared_master_rhea_reaction"
  )),
  identical(
    expanded$microcompass$params$target_scope,
    "direct_kegg_reactome_master_rhea_noncore_only"
  ),
  expanded$microcompass$params$structural_model_reused_exactly,
  all(expanded$microcompass$model_cache_summary$reused_without_rebuilding)
)
```

The selected core reactions are mapping anchors and are not scored again. Only non-core reactions directly sharing a KEGG, Reactome, or master-Rhea identifier with a selected core enter the second LP. Same-subsystem and transitive expansion are excluded. The exact original union-GEM files and bounds are reused.

## Assemble and interpret results

```{r}
result <- rc_regcompass_step_results(
  grn = step1,
  metacells = step2,
  meta_modules = step3,
  layer1 = step4,
  layer2 = step5,
  gem = gem,
  outdir = "steps/06_results"
)

stopifnot(
  identical(result$version, "1.8.2"),
  nrow(result$reaction_ranking) > 0,
  nrow(result$reaction_catalog) > 0,
  nrow(result$reaction_evidence) > 0
)
```

For condition tests and plots, see `docs/condition-reaction-statistics.md`. For medium definitions, see `docs/medium-presets.md`. For restart and diagnostics, see `docs/tutorial-03-advanced-restart.md`.

Metacell-level condition comparisons describe separation within the analyzed dataset; they are not automatically biological-replicate-level treatment inference.
