---
title: "RegCompassR 2.0.0 condition-comparable Pando workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RegCompassR 2.0.0 condition-comparable Pando workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

# Overview

RegCompassR converts paired single-cell RNA+ATAC evidence into reaction-level
support and directional COMPASS-like scores.

```text
one Pando motif/domain edge dictionary per cell type
→ pooled TF-RNA × peak-ATAC predictor transforms
→ condition-sparse selection followed by common-metric refit
→ explicit condition-versus-reference comparison mask
→ supported metabolic genes and complete-GPR cores
→ one ordered biological expansion pass
→ condition × broad cell type SuperCells
→ integrated reaction support and penalties
→ one shared medium-specific GEM
→ directional LP scoring
```

Pando 1.4.0 is the sole GRN estimator. RegCompass consumes its
`ConditionGRNFit v4` contracts without refitting coefficient matrices.

# Prepare the human GEM and medium

```{r}
library(RegCompassR)
library(BSgenome.Hsapiens.UCSC.hg38)

gem <- rc_prepare_gem(
  species = "human",
  version = "2.0.0",
  source = "bundled"
)

medium_scenarios <- rc_make_medium_scenarios(
  gem,
  scenario = "high_glucose",
  species = "human"
)
```

When `pfm` is omitted, RegCompass loads Pando's bundled `motifs` object. Human
analyses also use the bundled hg38 phastCons plus SCREEN ccRE union unless
`pando_initiate_args$regions` is supplied.

# Complete workflow

```{r}
result <- rc_run_regcompass_one_shot(
  object = A,
  outdir = "RegCompass_result",
  genome = BSgenome.Hsapiens.UCSC.hg38,
  species = "human",
  gem = gem,
  condition_col = "Group",
  celltype_col = "cell_type",
  sample_col = "sample_id",

  pando_args = list(
    min_cells = 100L,
    min_abs_estimate = 0,
    min_model_rsq = 0.1,
    pando_infer_args = list(
      method = "shared_baseline_condition_sparse",
      candidate_screen = "motif_domain",
      tf_cor = 0.1,
      peak_cor = 0,
      alpha = 0.5,
      condition_mix = 0.5,
      condition_weight = "equal",
      cv_block_col = "sample_id",
      reference_condition = "Control",
      nlambda = 50L,
      nfolds = 5L,
      lambda_selection = "lambda.1se",
      scale = TRUE
    )
  ),

  fragment_files = FALSE,
  metacell_args = list(
    rna_reduction = "pca",
    rna_dims = 1:30,
    atac_reduction = "lsi",
    atac_dims = 2:30,
    gamma = 30L,
    seed = 12345L,
    min_cells_per_stratum = 500L,
    min_metacell_size = 10L,
    min_metacells_per_stratum = 2L,
    overwrite = FALSE
  ),

  layer1_args = list(
    comparison_support = "auto",
  regulatory_alpha = 1,
    gpr_and_method = "min"
  ),
  medium_scenarios = medium_scenarios,
  model_mode = "meta_module_gem",
  layer2_args = list(
    target_direction = "both",
    solver = "highs",
    model_params = list(
      completion_time_limit = 600,
      fastcore_epsilon = 1e-4,
      max_support_reactions = 2000,
      strict = TRUE
    )
  ),
  upstream_workers = 6L,
  layer2_workers = 30L
)
```

The one-shot runner controls Pando execution through `upstream_workers` and a
stage-scoped backend. `pando_infer_args$parallel` and `BPPARAM` are not part of
the nested inference example.

# Pando shared-design contract

For edge `e` and unit `u`, Pando fits the interaction predictor:

```text
x[e, u] = RNA[TF(e), u] × ATAC[peak(e), u]
```

The final interaction is centered and scaled once over all cells of the cell
type. Conditions share this transform, the target transform, and a
target-specific lambda path/selection. With
`method = "shared_baseline_condition_sparse"`, `condition_mix = 0.5`, and
`condition_weight = "equal"`, condition-specific supports are selected jointly and then refit on one pooled common metric.

The interaction-safe candidate default is `motif_domain`. The optional `pooled_within_condition` mode applies response-dependent marginal-correlation screening and is a sensitivity mode rather than a canonical default; its OOF score is not used as confirmatory Layer 1 reliability (`q = 0`).

# Reference contrasts and comparison support

Pando exports:

```text
contrast[e, c] = beta[e, c] - beta[e, reference]
comparison_mask[e, c] =
  eligibility_mask[e, c] && eligibility_mask[e, reference]
```

RegCompass requires the explicit `comparison_mask`. A coefficient fixed to zero
because an edge is non-estimable cannot enter the active effect table or the
Layer 1 penalty projection.

```{r}
all_effects <- result$grn$tf_peak_gene_condition_effect_all
all_effects[, c(
  "edge_id", "Group", "cell_type",
  "condition_estimate", "reference_estimate", "condition_effect",
  "comparable_to_reference"
)]

table(all_effects$comparable_to_reference, useNA = "ifany")
```

# Pando argument routing

```text
pando_initiate_args → Pando::initiate_grn()
pando_motif_args    → Pando::find_motifs()
pando_infer_args    → Pando::infer_condition_grn()
```

RegCompass owns the object, assays, motif object, genome, metadata columns,
complete GEM target set, network name, minimum condition size, error policy,
and `BPPARAM`. Nested attempts to override these fields stop before Pando is
called. Stage 1 also rejects Pando aggregation columns because Stage 2 owns
metacell construction.

# Restartable Stage 1

```{r}
library(BiocParallel)

upstream_bp <- if (.Platform$OS.type == "windows") {
  SnowParam(workers = 6L, type = "SOCK")
} else {
  MulticoreParam(workers = 6L)
}

step1 <- rc_regcompass_step_grn(
  object = A,
  gem = gem,
  outdir = "RegCompass_steps/01_grn",
  genome = BSgenome.Hsapiens.UCSC.hg38,
  species = "human",
  condition_col = "Group",
  celltype_col = "cell_type",
  pando_args = list(
    min_cells = 100L,
    min_model_rsq = 0.1,
    pando_infer_args = list(
      candidate_screen = "motif_domain",
      reference_condition = "Control"
    )
  ),
  parallel = TRUE,
  BPPARAM = upstream_bp
)

step1$params$pando_parallel
step1$grn_result$condition_grn_fits
step1$grn_result$normalization_policy[c(
  "pando_motifs",
  "pando_regions",
  "pando_candidate_screen",
  "comparison_support",
  "pando_peak_cor",
  "min_model_rsq"
)]
```

With `parallel = TRUE` and `BPPARAM = NULL`, RegCompass enables Pando's native
map backend. With `parallel = FALSE`, Stage 1 is serial. `BPPARAM = TRUE` is
invalid.

# Stages 2–6

```{r}
step2 <- rc_regcompass_step_metacells(
  object = A,
  outdir = "RegCompass_steps/02_metacells",
  condition_col = "Group",
  celltype_col = "cell_type",
  fragment_files = FALSE,
  metacell_args = list(
    rna_reduction = "pca",
    rna_dims = 1:30,
    atac_reduction = "lsi",
    atac_dims = 2:30,
    gamma = 30L,
    seed = 12345L,
    min_cells_per_stratum = 500L,
    min_metacell_size = 10L,
    min_metacells_per_stratum = 2L
  )
)

step3 <- rc_regcompass_step_meta_modules(
  grn = step1,
  metacells = step2,
  gem = gem,
  outdir = "RegCompass_steps/03_meta_modules"
)

step4 <- rc_regcompass_step_layer1(
  grn = step1,
  metacells = step2,
  meta_modules = step3,
  gem = gem,
  outdir = "RegCompass_steps/04_layer1",
  regulatory_alpha = 1,
  gpr_and_method = "min",
  parallel = TRUE,
  BPPARAM = upstream_bp
)

step5 <- rc_regcompass_step_layer2(
  layer1 = step4,
  meta_modules = step3,
  gem = gem,
  medium_scenarios = medium_scenarios,
  outdir = "RegCompass_steps/05_layer2",
  model_mode = "meta_module_gem",
  layer2_args = list(
    target_direction = "both",
    solver = "highs",
    model_params = list(
      completion_time_limit = 600,
      fastcore_epsilon = 1e-4,
      max_support_reactions = 2000,
      strict = TRUE
    )
  )
)

result <- rc_regcompass_step_results(
  grn = step1,
  metacells = step2,
  meta_modules = step3,
  layer1 = step4,
  layer2 = step5,
  gem = gem,
  outdir = "RegCompass_steps/06_results",
  species = "human"
)
```

Stage 3 performs one complete-GPR/subsystem/cross-reference expansion pass and
does not run FASTCORE. Stage 4 delegates cell-first TF×ATAC projection to Pando and aggregates completed gene projections through exact SuperCell membership. Stage 5 performs the only global FASTCORE
completion for the shared medium-specific model.

# Mouse input

Pando's bundled regulatory-region objects are hg38. Mouse runs must supply a
build-matched `GRanges` object:

```{r}
library(BSgenome.Mmusculus.UCSC.mm10)

mouse_regions <- readRDS("mm10_regulatory_regions.rds")

mouse_step1 <- rc_regcompass_step_grn(
  object = A_mouse,
  gem = mouse_gem,
  outdir = "RegCompass_mouse/01_grn",
  genome = BSgenome.Mmusculus.UCSC.mm10,
  species = "mouse",
  condition_col = "Group",
  celltype_col = "cell_type",
  pando_args = list(
    pando_initiate_args = list(regions = mouse_regions),
    pando_infer_args = list(
      candidate_screen = "motif_domain",
      reference_condition = "Control"
    )
  )
)
```

RegCompass stops instead of silently applying hg38 regions to mouse ATAC peaks.


# Single-sample conditions and SuperCell strata

SuperCell hard strata are exactly condition × broad cell type; sample IDs are diagnostics only. If a condition has one biological sample, cell-level folds select lambda, sample-blocked OOF is unavailable, and Layer 1 uses regulatory reliability zero rather than treating metacells as replicates.
