---
title: "RegCompassR: from multiome input to directional scoring"
author: "RegCompass Developers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RegCompassR: from multiome input to directional scoring}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  purl = FALSE
)
```

## Workflow contract

RegCompassR exposes six supported functions. The explicit path prepares a
species GEM model with `rc_prepare_gem()`, defines exchange constraints
with `rc_make_medium_scenarios()`, and passes both to `rc_run_regcompass()`.
`rc_run_regcompass_one_shot()` performs those setup calls when `gem` or
`medium_scenarios` is omitted.

The integrated calculation follows this order:

```text
Seurat RNA + ATAC object plus optional fragment files
  -> condition x sample x cell-type strata
  -> fixed-resolution metacells within each stratum
  -> fragment re-aggregation when fragment files are supplied; otherwise object ATAC peak counts
  -> stratum-specific Pando regulatory evidence
  -> conservative GPR reaction evidence
  -> local FASTCORE-completed reaction meta-modules
  -> sample-balanced Q95 diagnostics and one shared GEM
  -> sample x cell-type directional minimum-discordance scoring
```

The primary score is an optimization penalty: lower values mean that the model
requires less disagreement with the observed evidence. It is not a measured
flux, a probability, or an independent observation for every metacell.

With `sample_balance = TRUE`, global diagnostic weights are
\(w_{si}=1/(S n_s)\). Within a Q95 stratum they are recomputed as
\(w_{sci}=1/(S_c n_{sc})\), so each represented biological sample has equal
total mass in that stratum. These weights affect Q95 and relative-state
diagnostics only; they do not multiply the zero-preserving absolute expression
support. Biological inference is balanced by the default `sample_celltype`
unit, which yields one unit per sample and represented cell type.

## 1. Validate the multiome input

`object` must inherit from `Seurat`, contain RNA and chromatin assays, and have
non-missing metadata columns for biological sample, condition, and cell type.
If matching fragment files are available, cell barcodes must agree across the
object and fragment files. Fragment files must be bgzip-compressed and
tabix-indexed. They may already be registered on the chromatin assay or be
supplied explicitly. If no matching fragment files are available, set
`fragment_files = FALSE`; RegCompass skips fragment aggregation and uses ATAC
peak raw counts already stored in the object.

```{r input-contract, purl=FALSE}
required_metadata <- c("sample_id", "condition", "cell_type")
stopifnot(inherits(object, "Seurat"))
stopifnot(all(required_metadata %in% colnames(object[[]])))
stopifnot(all(c("RNA", "ATAC") %in% names(object@assays)))
stopifnot(!anyNA(object[[]][, required_metadata]))

# A sample-aware manifest is the least ambiguous representation when fragments exist.
fragment_files <- data.frame(
  sample_id = c("donor_01", "donor_02"),
  assay = "ATAC",
  fragment_file = c(
    "fragments/donor_01.fragments.tsv.gz",
    "fragments/donor_02.fragments.tsv.gz"
  )
)
stopifnot(all(file.exists(fragment_files$fragment_file)))
stopifnot(all(file.exists(paste0(fragment_files$fragment_file, ".tbi"))))

with(object[[]], table(condition, sample_id, cell_type))
```

Every biological sample must contribute at least one retained stratum. Choose
`min_cells_per_stratum`, metacell resolution (`gamma`), and Pando's minimum
metacell count together; otherwise the global completion barrier will stop the
run rather than silently discard a sample.

## 2. Prepare and inspect the GEM

Pin a Human-GEM 2 release for reproducibility. The helper downloads, converts,
annotates, validates, and caches the model.

```{r gem, purl=FALSE}
library(RegCompassR)

human2_gem <- rc_prepare_human2_gem(
  version = "2.0.0",
  cache_dir = "model-cache"
)
# For mouse data, use rc_prepare_mouse_gem(version = "1.8.0").

dim(human2_gem$S)
head(human2_gem$gpr_table)
table(human2_gem$reaction_meta$role, useNA = "ifany")
human2_gem$model_info[c("source", "version", "commit", "checksum")]
```

Do not use `version = "latest"` for a reproducible analysis. If it is required
for exploration, it must be explicitly enabled with `allow_latest = TRUE`.

## 3. Define one shared medium

The README quick start uses the published `normal_human_plasma` background.
`compass_model_bounds` remains available as a technical model-defined
environment. Neither should be interpreted as measured sample-specific
metabolite concentrations.

Preset scenarios can also be returned together with a user-defined `"custom"`
scenario in one medium table. Current optional named backgrounds are
`normal_human_plasma`, `mouse_plasma`, `rpmi1640`, `dmem_high_glucose`,
`low_glucose`, `high_glucose`, `low_lactate`, `high_lactate`,
`low_glutamine`, `minimal`, and `permissive_all_exchange`. Removed
compatibility aliases such as `human_plasma`, `hplm`, `blood_like`,
`culture_like`, `tumor_low_glucose`, and `lactate_available` are intentionally
unsupported; use canonical preset names or a custom medium table.

```{r medium, purl=FALSE}
medium <- rc_make_medium_scenarios(
  gem = human2_gem,
  scenario = "normal_human_plasma"
)

stopifnot(all(medium$exchange_reaction_id %in% colnames(human2_gem$S)))
stopifnot(all(medium$lb <= medium$ub))
stopifnot(identical(unique(medium$condition), "all"))
head(medium[c("exchange_reaction_id", "lb", "ub", "available")])
```

Use a custom table only when bounds have an external justification. It must
include `medium_scenario_id`, `exchange_reaction_id`, `lb`, `ub`, and
`available`. Keep the medium shared across conditions unless the experimental
design independently establishes condition-specific exchange constraints.

## 4. Run the explicit workflow

The explicit entry point makes every major input visible. The PFM collection,
candidate regulatory regions, and genome object shown below match the current
human quick-start example; use organism- and genome-build-matched objects for
other analyses.

```{r explicit-run, purl=FALSE}
library(Pando)
library(BSgenome.Hsapiens.UCSC.hg38)
data(motifs, package = "Pando")
data(SCREEN.ccRE.UCSC.hg38, package = "Pando")

result <- rc_run_regcompass(
  object = object,
  gem = human2_gem,
  outdir = "RegCompass_result",
  pfm = motifs,
  genome = BSgenome.Hsapiens.UCSC.hg38,
  fragment_files = FALSE,  # use object ATAC peak counts; pass fragment_files to re-aggregate
  sample_col = "sample_id",
  condition_col = "condition",
  celltype_col = "cell_type",
  rna_assay = "RNA",
  atac_assay = "ATAC",
  model_mode = "meta_module_gem",
  medium_scenarios = medium,
  metacell_args = list(
    gamma = 150,
    min_cells_per_stratum = 500,
    min_metacell_size = 10
  ),
  pando_args = list(
    min_metacells = 10,
    pando_initiate_args = list(
      regions = SCREEN.ccRE.UCSC.hg38
    ),
    pando_infer_args = list(
      method = "glm",
      tf_cor = 0.1,
      peak_cor = 0.05,
      adjust_method = "fdr"
    )
  ),
  layer1_args = list(
    local_fastcore = TRUE,
    sample_balance = TRUE,
    expression_batch_correction = "none"
  ),
  layer2_args = list(
    target_direction = "both",
    solver = "highs"
  ),
  inference_unit = "sample_celltype",
  upstream_workers = 4,
  layer2_workers = 4,
  parallel_backend = "snow"
)
```

The four `*_args` values are lists by contract. Integrated fields such as
`layer1`, `gem`, `unit`, and `medium_scenarios` are owned by the workflow and
must not be overridden through `layer2_args`. Use `inference_unit` on
`rc_run_regcompass()` to select the Layer 2 unit. Candidate regions belong in
`pando_args$pando_initiate_args$regions`, matching `Pando::initiate_grn()`.

## 5. Equivalent one-shot entry point

For a first run, omit `gem` and `medium_scenarios`; the one-shot function
constructs them and delegates to the same integrated workflow. The default is
`species = "human"`; set `species = "mouse"` to automatically prepare
Mouse-GEM 1.8.0 and the mouse physiological medium. Arguments in `...` are
passed to `rc_run_regcompass()`.

```{r one-shot, purl=FALSE}
result <- rc_run_regcompass_one_shot(
  object = object,
  outdir = "RegCompass_result",
  pfm = motifs,
  genome = BSgenome.Hsapiens.UCSC.hg38,
  fragment_files = FALSE,  # use existing ATAC peak counts; pass paths to re-aggregate fragments
  species = "human",
  gem_version = "2.0.0",
  medium_scenario = "normal_human_plasma",
  sample_col = "sample_id",
  condition_col = "condition",
  celltype_col = "cell_type",
  metacell_args = list(
    gamma = 150,
    min_cells_per_stratum = 500,
    min_metacell_size = 10
  ),
  pando_args = list(
    min_metacells = 10,
    pando_initiate_args = list(
      regions = SCREEN.ccRE.UCSC.hg38
    ),
    pando_infer_args = list(
      method = "glm",
      tf_cor = 0.1,
      peak_cor = 0.05,
      adjust_method = "fdr"
    )
  ),
  layer1_args = list(
    local_fastcore = TRUE,
    sample_balance = TRUE,
    expression_batch_correction = "none"
  ),
  layer2_args = list(
    target_direction = "both",
    solver = "highs"
  )
)
```

Set `fragment_files = FALSE` in the one-shot call when no matching fragment file
is available; pass the fragment manifest/path when fragments should be
re-aggregated.
