---
title: "RegCompassR workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RegCompassR workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

# Overview

RegCompassR converts paired single-cell RNA+ATAC data into regulatory reaction
support and directional metabolic scores.

```text
>=2 condition levels → common-dictionary condition GRNs
0 or 1 condition level → original per-cell-type Pando GRN
```

For multiple conditions, each broad cell type uses one two-stage Pando design:

```text
pooled candidate discovery
+ condition-specific candidate discovery
→ exact TF-peak-target union
→ frozen target-specific dictionary
→ one unscaled Gaussian identity GLM per condition
→ estimable BH padj < 0.05 condition effects
```

The first-stage candidate fits are not final effects. No pooled coefficient is
used to rescale a condition coefficient. Conditions may have different
significant edges and opposite directions while retaining the same predictor
definition and numerical scale.

Stage 2 calls `SuperCell::SCimplify_by_graph_group()` with one independent
multimodal WNN graph per broad cell type. All conditions of that type jointly
determine the graph; condition splits parent membership only after clustering.
Final metacells are condition-pure without condition-specific graphs.

Stage 3 builds condition-specific biological meta-modules, then unions conditions
within the same cell type only. Stage 5 constructs one union GEM for every
`cell_type × medium_scenario` pair and runs FASTCORE independently in each model.
Conditions of one cell type share that cell-type model; different cell types do
not share reaction unions, FASTCORE support, model files or directional `vmax`.

# Medium and GEM

```{r}
library(RegCompassR)
library(BSgenome.Hsapiens.UCSC.hg38)

gem <- rc_prepare_gem(
  species = "human",
  version = "2.0.0",
  source = "bundled"
)

medium_scenarios <- rc_make_medium_scenarios(
  gem,
  scenario = c("normal_human_plasma", "high_glucose", "low_glucose"),
  species = "human"
)
```

Built-in scenarios can be combined with custom reaction-level bounds or
metabolite-level availability. Medium definitions change exchange constraints;
they are not measured transporter fluxes.

# One-shot run

```{r}
result <- rc_run_regcompass_one_shot(
  object = A,
  gem = gem,
  outdir = "RegCompass_result",
  genome = BSgenome.Hsapiens.UCSC.hg38,
  species = "human",
  condition_col = "Group",
  celltype_col = "cell_type",
  pando_args = list(
    min_cells = 300L,
    pando_infer_args = list(
      tf_cor = 0.1,
      peak_cor = 0,
      adjust_method = "BH",
      padj_threshold = 0.05,
      rank_action = "mark",
      min_residual_df = 1L
    )
  ),
  metacell_args = list(
    rna_reduction = "pca",
    rna_dims = 1:30,
    atac_reduction = "lsi",
    atac_dims = 2:30,
    gamma = 30L,
    k.knn = 30L,
    seed = 12345L
  ),
  layer1_args = list(
    gpr_and_method = "min"
  ),
  medium_scenarios = medium_scenarios,
  model_mode = "meta_module_gem",
  layer2_args = list(
    target_direction = "both",
    solver = "highs"
  ),
  upstream_workers = 6L,
  layer2_workers = 30L
)
```

Stage 1 fixes `min_cells` at 300. `upstream_workers` controls GRN and other
upstream tasks. `layer2_workers` controls independent cell-type/medium model
construction, cell-type FASTCORE completion, directional `vmax` calculation and
matching-metacell LP tasks. Layer 2 workers force HiGHS and numerical libraries
to one internal thread to avoid nested oversubscription.

The following retired condition-GRN parameters are rejected:

```text
candidate_screen
condition_mix
condition_weight
alpha
nlambda / lambda / lambda_min_ratio
outer_nfolds / inner_nfolds
lambda_selection
scale
engine_control
comparison_conditions
```

# Restartable stages

The same workflow can be run as six persisted stages:

```{r}
step1 <- rc_regcompass_step_grn(...)
step2 <- rc_regcompass_step_metacells(..., grn = step1)
step3 <- rc_regcompass_step_meta_modules(
  grn = step1, metacells = step2, gem = gem, outdir = "03_meta_modules"
)
step4 <- rc_regcompass_step_layer1(
  grn = step1, metacells = step2, meta_modules = step3,
  gem = gem, outdir = "04_layer1"
)
step5 <- rc_regcompass_step_layer2(
  layer1 = step4, meta_modules = step3, gem = gem,
  medium_scenarios = medium_scenarios, outdir = "05_layer2"
)
result <- rc_regcompass_step_results(
  grn = step1, metacells = step2, meta_modules = step3,
  layer1 = step4, layer2 = step5, gem = gem,
  outdir = "06_results"
)
```

Each stage writes an RDS artifact plus progress and timing logs. Changes to only
the medium or LP controls require rerunning Stage 5 and final results, not the
GRN and metacell stages. Changes to GPR rules or biological reaction annotations
require rerunning Stage 3 onward because they alter each cell type's catalogue.

# Condition effect and penalty

For an estimable edge, `estimate` is the complete fixed-dictionary condition
coefficient. Pando defines:

```text
penalty_effect = estimate when padj < 0.05
penalty_effect = 0 otherwise
```

The multi-condition handoff applies no additional absolute-effect or model-R²
standard-Pando fallback. The complete coefficient table is retained even when
an edge does not enter the penalty. A zero-variance, aliased, non-finite, or
insufficient-residual-df coefficient remains `NA`.

The GLM P value is conditional on the frozen candidate dictionary and does not
include selective-inference correction for candidate discovery.

# Cell-first projection

RegCompass reconstructs each paired-cell predictor `TF RNA × peak ATAC`, applies
the cell's condition-specific `penalty_effect`, and sums edge contributions by
target. It then aggregates target scores by exact SuperCell membership. It does
not recompute TF×ATAC from metacell averages or refit after aggregation.

```{r}
result$grn$tf_peak_gene_condition_effect_all
result$grn$tf_peak_gene_condition_effect
result$layer1$gene_regulatory_modifier
result$microcompass$penalty
result$reaction_ranking
result$condition_contrast
```

for backward compatibility. The primary and common fields are aliases of the
fixed-dictionary projection, and the condition-unique compatibility matrix is
zero. These names do not imply OOF fitting or a common-support model.

# Cell-type union GEMs

Inspect the structural model catalogue with:

```{r}
result$microcompass$model_cache_summary[, c(
  "cell_type",
  "medium_scenario",
  "file",
  "file_checksum",
  "n_celltype_biological_reactions",
  "n_celltype_fastcore_support_reactions"
)]
```

The unique structural key is `cell_type × medium_scenario`. Within that key, all
conditions use the same stoichiometric model, bounds and directional `vmax`.
Rows from another cell type are absent from the comparison rather than evaluated
on a common global GEM.

# No condition or one condition

With no usable condition column or one observed condition level, RegCompass runs
original Pando independently within each retained broad cell type. It does not
create condition coefficients or a synthetic condition fit contract. Structural
meta-modules and union GEMs remain separated by cell type.

# Optional targeted reaction remapping

After a completed `model_mode = "meta_module_gem"` run, directly database-linked
reactions can be scored without rebuilding the cached model:

```{r}
targeted <- rc_regcompass_step_target_union(
  layer1 = result$layer1,
  meta_modules = result$meta_modules,
  layer2 = result$microcompass,
  gem = gem,
  outdir = "RegCompass_targeted",
  core_reaction_ids = c("MAR04381", "MAR04379"),
  layer2_args = list(target_direction = "both", solver = "highs")
)
```

The second pass reuses the exact cached union GEMs for the corresponding cell
type and medium. Candidate reactions are intersected across media within that
cell type. FASTCORE is not rerun, and cached models from different cell types are
never merged.

# Documentation

- [Tutorial 1: one-shot workflow](../docs/tutorial-01-quick-start.md)
- [Tutorial 2: stepwise workflow](../docs/tutorial-02-stepwise-audit.md)
- [Tutorial 3: mathematical model](../docs/tutorial-03-mathematical-model.md)
- [Tutorial 4: targeted reaction remapping](../docs/tutorial-04-targeted-reaction-remapping.md)
- [Tutorial 5: condition comparison](../docs/tutorial-05-condition-differential-analysis.md)
- [Medium scenarios and evidence](../docs/medium-presets.md)
- [Stage contracts](../docs/stage-interface-contracts.md)
- [Public API](../docs/functions.md)
