---
title: "RegCompassR workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RegCompassR workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

# Overview

RegCompassR converts paired single-cell RNA+ATAC data into regulatory reaction
support and directional metabolic scores.

Each stage writes `step_timing.tsv` and a structured `step_progress.tsv`. With `progress = TRUE`, the console also reports phase, percent and elapsed time. Stage 1 additionally records cell types, conditions, metabolic targets, outer/inner folds, lambda path size, hybrid solver route, contract extraction and artifact writing. Setting `progress = FALSE` suppresses console output but retains the audit log.

```text
>=2 condition levels -> condition_grn
0 or 1 condition level -> standard_pando using Pando::infer_grn()
```

Condition mode uses condition-full outer-heldout projection as the primary
regulatory route. Jointly estimable edges form the common-support component. A
non-estimable edge side remains projectable and contributes exactly zero.

Stage 2 calls `SuperCell::SCimplify_by_graph_group()` with:

```text
cell.graph.group = broad cell type
cell.split.condition = condition
gamma = 30L
k.knn = 30L
```

Each broad cell type has one independent native RNA+ATAC WNN graph. All
conditions within that cell type jointly determine adaptive modality weights,
neighbours and Walktrap parent clusters. Condition splits parent membership only
after clustering, so final metacells are condition-pure without fitting
condition-specific graphs. No sample or temporary condition-by-cell-type stratum
is used.

The Stage 2 provenance contract records:

```text
one_independent_WNN_graph_per_cell_type
all_conditions_joint_within_cell_type_graph
after_joint_WNN_graph_clustering
adaptive_WNN_within_cell_type
temporary_combined_stratum = FALSE
```

Mathematical definitions are centralized in
[Tutorial 3](../docs/tutorial-03-mathematical-model.md).

# Run

```{r}
library(RegCompassR)
library(BSgenome.Hsapiens.UCSC.hg38)

gem <- rc_prepare_gem(
  species = "human",
  version = "2.0.0",
  source = "bundled"
)

medium_scenarios <- rc_make_medium_scenarios(
  gem,
  scenario = "normal_human_plasma",
  species = "human"
)

result <- rc_run_regcompass(
  object = A,
  gem = gem,
  outdir = "RegCompass_result",
  genome = BSgenome.Hsapiens.UCSC.hg38,
  species = "human",
  condition_col = "Group",
  celltype_col = "cell_type",
  pando_args = list(
    min_cells = 100L,
    min_model_rsq = 0.1,
    pando_infer_args = list(
      candidate_screen = "motif_domain",
      condition_mix = 0.5,
      condition_weight = "equal",
      outer_nfolds = 5L,
      inner_nfolds = 5L,
      scale = TRUE
    )
  ),
  metacell_args = list(
    rna_reduction = "pca",
    rna_dims = 1:30,
    atac_reduction = "lsi",
    atac_dims = 2:30,
    gamma = 30L,
    k.knn = 30L,
    seed = 12345L
  ),
  layer1_args = list(
    comparison_support = "auto",
    regulatory_alpha = 1,
    gpr_and_method = "min"
  ),
  medium_scenarios = medium_scenarios,
  model_mode = "meta_module_gem"
)
```

Supported biological medium scenarios are:

```text
normal_human_plasma
mouse_plasma
high_glucose
low_glucose
high_lactate
low_lactate
low_glutamine
custom
```

Human composition is anchored to HPLM from *Cell* 2017 and updated HPLM from
*Cell Metabolism* 2021. Plasmax from *Science Advances* 2019 is retained as an
independent validation source and is not numerically averaged with HPLM. Mouse
composition follows the package's documented mouse-plasma evidence table.

All five challenge scenarios use
`medium_background_id = authoritative_HPLM_2017_2021` and override only the
named glucose, lactate, or glutamine concentration. Their output separates
`background_reference_doi`, `background_validation_reference_doi`, and
`challenge_reference_doi`, and records
`authoritative_HPLM_background_plus_named_nutrient_override` in
`scenario_construction`. User-defined `custom_medium` and `custom_metabolites`
remain supported through `scenario = "custom"` or `scenario = NULL`.

# Inspect

```{r}
result$analysis_mode
result$metacells$input_design
result$layer1$gene_projection_condition_full_oof
result$layer1$gene_projection_common_oof
result$microcompass$penalty_condition_full_oof
result$microcompass$penalty_common_oof
result$microcompass$penalty_condition_unique_increment
result$reaction_ranking
result$condition_contrast
```

`reaction_ranking` and `condition_contrast` use the condition-full penalty.
Common-support and RNA-only routes are decomposition/control outputs. All routes
reuse the exact same medium-specific model and `vmax`.

The canonical schema does not contain depth matching, common-depth restriction,
alpha sensitivity, zero-support sensitivity, or link-saturation propagation.

Metacells are within-dataset statistical units, not biological replicates.

# Optional targeted reaction remapping

After a completed `model_mode = "meta_module_gem"` run, directly database-linked
non-core reactions can be scored without rebuilding the model:

```{r}
targeted <- rc_regcompass_step_target_union(
  layer1 = result$layer1,
  meta_modules = result$meta_modules,
  layer2 = result$microcompass,
  gem = gem,
  outdir = "RegCompass_targeted",
  core_reaction_ids = c("MAR04381", "MAR04379"),
  layer2_args = list(target_direction = "both", solver = "highs")
)
```

The second pass reuses the exact Stage 5 union GEM and global FASTCORE
completion. In condition mode, `result$layer1$reaction_expression` is the
canonical alias of `reaction_expression_condition_full_oof`, so targeted targets
use the same primary evidence route as the original Stage 5 targets.

# Documentation

- [Tutorial 1](../docs/tutorial-01-quick-start.md)
- [Tutorial 2](../docs/tutorial-02-stepwise-audit.md)
- [Tutorial 3](../docs/tutorial-03-mathematical-model.md)
- [Tutorial 4: targeted reaction remapping](../docs/tutorial-04-targeted-reaction-remapping.md)
- [Tutorial 5: condition comparison](../docs/tutorial-05-condition-differential-analysis.md)
- [Medium scenarios and evidence](../docs/medium-presets.md)
- [Stage contracts](../docs/stage-interface-contracts.md)
- [Public API](../docs/functions.md)
