---
title: "RegCompassR 1.8.8 bootstrap-stable condition-sub-GRN workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RegCompassR 1.8.8 bootstrap-stable condition-sub-GRN workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

# Overview

RegCompassR 1.8.8 separates the shared regulatory hypothesis from condition-specific regulatory states.

```text
one validated Pando TF–peak–target universe per cell type
→ global backbone + symmetric condition deviations
→ full-size condition-stratified bootstrap stability
→ condition-specific metabolic target genes
→ complete-GPR condition cores
→ one ordered annotation expansion pass
→ one merged reaction catalogue
→ one medium-specific union GEM and one global FASTCORE completion
→ identical stoichiometric structure across conditions
→ condition/metacell-specific penalties and directional LP scores
```

For edge `e`,

\[
\theta_{e,c}=\beta_e+\delta_{e,c},
\qquad
\sum_c\delta_{e,c}=0.
\]

Every condition uses the same Pando candidate dictionary and contributes the same total regression loss weight. The canonical public workflow uses condition and cell type metadata only; it does not accept a biological-sample column.

# Prepare the GEM and medium

```{r}
library(RegCompassR)
library(BSgenome.Hsapiens.UCSC.hg38)

gem <- rc_prepare_gem(
  species = "human",
  version = "2.0.0",
  source = "bundled"
)

medium_scenarios <- rc_make_medium_scenarios(
  gem,
  scenario = "high_glucose",
  species = "human"
)
```

When `pfm` is omitted, RegCompass loads `data("motifs", package = "Pando")`. Human analyses use phastCons plus SCREEN ccRE regions; mouse analyses use phastCons regions unless overridden.

# Complete workflow

```{r}
result <- rc_run_regcompass_one_shot(
  object = A,
  outdir = "RegCompass_result",
  genome = BSgenome.Hsapiens.UCSC.hg38,
  species = "human",
  gem = gem,
  condition_col = "Group",
  celltype_col = "cell_type",

  grn_mode = "multitask_shared_backbone",
  pando_args = list(
    min_cells = 100,
    pando_design_args = list(
      peak_to_gene_method = "Signac",
      min_tf_detection = 0.01,
      min_peak_detection = 0.01,
      min_target_detection = 0.01
    )
  ),
  multitask_args = list(
    alpha = 0.5,
    global_penalty_factor = 1,
    deviation_penalty_factor = 2,
    lambda_rule = "lambda.1se",
    nfolds = 5,
    n_bootstrap = 100,
    min_selection_frequency = 0.7,
    min_sign_stability = 0.8,
    candidate_screen_threshold = 0,
    max_edges_per_target = Inf,
    seed = 12345L
  ),

  fragment_files = FALSE,
  metacell_args = list(
    rna_reduction = "pca",
    rna_dims = 1:30,
    atac_reduction = "lsi",
    atac_dims = 2:30,
    gamma = 30,
    seed = 12345L,
    min_cells_per_stratum = 500,
    min_metacell_size = 10,
    min_metacells_per_stratum = 2L,
    overwrite = FALSE
  ),

  layer1_args = list(
    regulatory_alpha = 1,
    gpr_and_method = "min"
  ),

  medium_scenarios = medium_scenarios,
  model_mode = "meta_module_gem",
  layer2_args = list(
    target_direction = "both",
    solver = "highs",
    model_params = list(
      completion_time_limit = 600,
      fastcore_epsilon = 1e-4,
      max_support_reactions = 2000,
      strict = TRUE
    )
  ),
  upstream_workers = 6,
  layer2_workers = 30
)
```

# Shared candidate background

`Pando::prepare_grn_design()` constructs structural candidates before fitting. An edge requires a target-linked regulatory region, a matched TF motif, an exact measured ATAC feature, and measured TF/target RNA features. Predictors are deduplicated by `(TF, ATAC feature, target)`, and the version-2 design fingerprint is validated before use.

The predictor is

\[
x_{e,u}=T_{t,u}A_{p,u}.
\]

The default candidate screen and edge cap preserve the complete structural universe. A positive screen threshold remains shared across conditions.

# Condition model and bootstrap stability

Target and predictor values are centred within condition. Shared scaling is calculated across all conditions of the cell type.

The symmetric condition matrix is

\[
H=I_C-\frac1C\mathbf 1\mathbf 1^T.
\]

Because the symmetric deviation columns are rank deficient, `alpha` must remain below one. The ridge component makes the regularised solution unique and reference-condition invariant.

For each bootstrap replicate, condition `c` is resampled with replacement at its original size `n_c`. The bootstrap target and predictors are re-centred within condition, scaled by the full-data shared scale, and fitted at the full-data selected lambda.

\[
\Pi_{e,c}=\frac1B\sum_bI(\theta^{(b)}_{e,c}\neq0),
\]

\[
\rho_{e,c}=\left|
\frac{\sum_bI(\theta^{(b)}_{e,c}\neq0)\operatorname{sign}(\theta^{(b)}_{e,c})}
{\sum_bI(\theta^{(b)}_{e,c}\neq0)}
\right|,
\]

\[
\widetilde\theta_{e,c}
=\widehat\theta_{e,c}\Pi_{e,c}\rho_{e,c}.
\]

Regularised coefficients do not receive classical adjusted p-values; `padj` remains `NA`. Bootstrap quantifies selection stability under cell resampling, not biological-replicate significance.

# Condition genes and complete-GPR cores

A gene enters the condition target set when at least one active bootstrap-stable edge targets it. Positive and negative directions both establish membership.

For reaction `r` with GPR branches `B_{r,k}`:

\[
Core_{r,c}=1
\iff
\exists k:B_{r,k}\subseteq G_c.
\]

Stage 3 then performs exactly one ordered expansion:

```text
complete-GPR cores
→ core-reaction subsystems
→ direct KEGG/Reactome equivalents
→ direct master-Rhea equivalents
→ stop
```

# Layer 1 regulatory projection

Stage 1 uses RNA and ATAC to fit the GRN. Stage 4 reprojects only metacell ATAC. TF edges sharing one measured peak are signed-summed before projection. One target denominator is shared across conditions, preventing condition-wise rescaling of weak and strong sub-GRNs to equal magnitude.

A gene with no active condition edge has `R = 0`, and the log-odds update returns RNA-only support exactly:

\[
C^{MO}
=
\frac{C^{RNA}2^{\alpha R}}
{1-C^{RNA}+C^{RNA}2^{\alpha R}}.
\]

# Shared metabolic model

Condition core sets may differ, but Stage 3 merges their reaction catalogues before Stage 5. For each medium, Stage 5 constructs one union GEM and runs the only global FASTCORE completion. All conditions and metacells reuse identical reaction IDs, stoichiometric matrix, and bounds.

`completion_time_limit` limits FASTCORE union-GEM construction only. Directional scoring LPs have no time-limit parameter.

# Stepwise inspection

```{r}
step1 <- rc_regcompass_step_grn(
  object = A,
  gem = gem,
  outdir = "RegCompass_steps/01_grn",
  genome = BSgenome.Hsapiens.UCSC.hg38,
  species = "human",
  condition_col = "Group",
  celltype_col = "cell_type",
  grn_mode = "multitask_shared_backbone",
  pando_args = list(
    min_cells = 100,
    pando_design_args = list(
      peak_to_gene_method = "Signac"
    )
  ),
  multitask_args = list(
    alpha = 0.5,
    n_bootstrap = 100,
    max_edges_per_target = Inf,
    seed = 12345L
  )
)

step2 <- rc_regcompass_step_metacells(
  object = A,
  outdir = "RegCompass_steps/02_metacells",
  condition_col = "Group",
  celltype_col = "cell_type",
  fragment_files = FALSE,
  metacell_args = list(
    rna_reduction = "pca",
    rna_dims = 1:30,
    atac_reduction = "lsi",
    atac_dims = 2:30,
    gamma = 30,
    seed = 12345L
  )
)

step3 <- rc_regcompass_step_meta_modules(
  grn = step1,
  metacells = step2,
  gem = gem,
  outdir = "RegCompass_steps/03_meta_modules"
)

step4 <- rc_regcompass_step_layer1(
  metacells = step2,
  meta_modules = step3,
  gem = gem,
  outdir = "RegCompass_steps/04_layer1",
  regulatory_alpha = 1,
  gpr_and_method = "min"
)

step5 <- rc_regcompass_step_layer2(
  layer1 = step4,
  meta_modules = step3,
  gem = gem,
  medium_scenarios = medium_scenarios,
  outdir = "RegCompass_steps/05_layer2",
  model_mode = "meta_module_gem",
  layer2_args = list(target_direction = "both", solver = "highs")
)

result <- rc_regcompass_step_results(
  grn = step1,
  metacells = step2,
  meta_modules = step3,
  layer1 = step4,
  layer2 = step5,
  gem = gem,
  outdir = "RegCompass_steps/06_results",
  species = "human"
)
```

Principal Stage 1 outputs:

```r
result$grn$tf_peak_gene_candidates
result$grn$tf_peak_gene_global
result$grn$tf_peak_gene_condition_all
result$grn$tf_peak_gene_significant
result$grn$condition_target_genes
result$grn$stability_diagnostics
result$grn$group_status
```

Condition and merged metabolic outputs:

```r
result$condition_grn_meta_modules$supported_metabolic_genes
result$condition_grn_meta_modules$core_gene_reaction
result$merged_grn_meta_modules$merged_core_reactions
result$merged_grn_meta_modules$merged_reaction_membership
result$microcompass$model_cache_summary
```

See `docs/multitask-shared-grn.md` for complete equations and object contracts.
