---
title: "RegCompassR 1.8.4 significant-Pando-target workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RegCompassR 1.8.4 significant-Pando-target workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

# Overview

RegCompassR converts paired single-cell RNA+ATAC evidence into reaction-level support and directional COMPASS-like scores.

```text
condition × cell type Pando TF–peak–GEM-gene models
→ significant metabolic target-gene sets
→ complete-GPR core reactions
→ one ordered subsystem/cross-reference expansion pass
→ merged meta-module catalogue
→ integrated RNA+ATAC reaction support
→ COMPASS-compatible GPR-AND aggregation
→ medium-specific union GEM
→ single global FASTCORE completion
→ directional LP scoring
```

Stage 3 maps significant targets to complete-GPR cores and applies one ordered annotation pass. Only the medium-constrained Stage 5 model is called a union GEM and runs FASTCORE completion.

# Prepare the GEM and medium

```{r}
library(RegCompassR)
library(BSgenome.Hsapiens.UCSC.hg38)

gem <- rc_prepare_gem(
  species = "human",
  version = "2.0.0",
  source = "bundled"
)

medium_scenarios <- rc_make_medium_scenarios(
  gem,
  scenario = "high_glucose",
  species = "human"
)
```

When `pfm` is omitted, RegCompass loads:

```r
data("motifs", package = "Pando")
pfm <- motifs
```

The resulting `motifs` object is passed to `Pando::find_motifs()`. A user-supplied `pfm` overrides this default.

The default Pando regions are species-specific:

```text
human: phastConsElements20Mammals.UCSC.hg38 ∪ SCREEN.ccRE.UCSC.hg38
mouse: phastConsElements20Mammals.UCSC.hg38 only
```

An explicit `pando_args$pando_initiate_args$regions` overrides either default.

# Complete workflow

Public runner arguments are ordered by processing sequence.

```{r}
result <- rc_run_regcompass_one_shot(
  object = A,
  outdir = "RegCompass_result",
  genome = BSgenome.Hsapiens.UCSC.hg38,
  species = "human",
  gem = gem,

  # Stage 1
  condition_col = "Group",
  celltype_col = "cell_type",
  pando_args = list(
    min_cells = 100,
    padj_threshold = 0.05,
    min_abs_estimate = 0,
    min_model_rsq = 0.1,
    require_padj = TRUE,
    pando_infer_args = list(
      method = "glm",
      tf_cor = 0.1,
      peak_cor = 0.01,
      adjust_method = "fdr",
      parallel = FALSE
    )
  ),

  # Stage 2
  sample_col = NULL,
  fragment_files = FALSE,
  metacell_args = list(
    rna_reduction = "pca",
    rna_dims = 1:30,
    atac_reduction = "lsi",
    atac_dims = 2:30,
    gamma = 30,
    seed = 12345L,
    min_cells_per_stratum = 500,
    min_metacell_size = 10,
    min_metacells_per_stratum = 2L,
    overwrite = FALSE
  ),

  # Stage 4
  layer1_args = list(
    regulatory_alpha = 1,
    gpr_and_method = "min"
  ),

  # Stage 5
  medium_scenarios = medium_scenarios,
  model_mode = "meta_module_gem",
  layer2_args = list(
    target_direction = "both",
    solver = "highs",
    model_params = list(
      completion_time_limit = 600,
      fastcore_epsilon = 1e-4,
      max_support_reactions = 2000,
      strict = TRUE
    )
  ),
  upstream_workers = 6,
  layer2_workers = 30
)
```

# Pando evidence filters

A TF–peak–target coefficient enters the significant evidence table only when all enabled filters pass:

- `padj_threshold = 0.05`: finite adjusted P value no greater than 0.05;
- `min_abs_estimate = 0`: finite absolute coefficient no smaller than the threshold;
- `min_model_rsq = 0.1`: finite target-model R² no smaller than 0.1;
- `require_padj = TRUE`: fail rather than silently proceed when adjusted P values are unavailable;
- `min_cells`: minimum cells required in the `condition × cell type` group.

Both positive and negative Pando coefficients count as regulatory evidence. The filters determine which GEM target genes enter Stage 3; coefficient direction is retained as provenance but does not change supported-gene membership.

# Metacell reductions, dimensions, and seed

The canonical Stage 2 defaults are:

```r
rna_reduction = "pca"
rna_dims = 1:30
atac_reduction = "lsi"
atac_dims = 2:30
gamma = 30L
seed = 12345L
min_cells_per_stratum = 100L
min_metacell_size = 20L
min_metacells_per_stratum = 2L
```

Reduction names must exist in `A@reductions`, and the requested dimensions must exist in their embedding matrices. LSI dimension 1 is excluded by default. For ordered condition strata, RegCompass passes the deterministic seed:

```text
seed_for_stratum = seed + stratum_index - 1
```

A precomputed Harmony embedding can replace PCA through `rna_reduction = "harmony"` and the corresponding `rna_dims`. Harmony changes membership geometry only; the original RNA and ATAC assay counts are aggregated. Changing cells, assays, reductions, dimensions, embedding values, seed, gamma, or thresholds invalidates the Stage 2 cache and requires `overwrite = TRUE`.

# Fixed Stage 3 expansion

Stage 3 expansion is fixed to one pass:

```text
core subsystem
→ KEGG/Reactome reaction equivalence
→ master-Rhea reaction equivalence
→ stop
```

# GPR-AND and model construction

`gpr_and_method` accepts `"min"`, `"median"`, or `"mean"` and defaults to `"min"`. FASTCORE is configured only at Stage 5 through `layer2_args$model_params`. `completion_time_limit` limits union-GEM construction only; the later scoring LPs have no time-limit parameter.

# Stepwise workflow

```{r}
step1 <- rc_regcompass_step_grn(
  object = A,
  gem = gem,
  outdir = "RegCompass_steps/01_grn",
  genome = BSgenome.Hsapiens.UCSC.hg38,
  species = "human",
  condition_col = "Group",
  celltype_col = "cell_type",
  pando_args = list(
    min_cells = 100,
    padj_threshold = 0.05,
    min_abs_estimate = 0,
    min_model_rsq = 0.1,
    require_padj = TRUE
  )
)

step2 <- rc_regcompass_step_metacells(
  object = A,
  outdir = "RegCompass_steps/02_metacells",
  sample_col = NULL,
  condition_col = "Group",
  celltype_col = "cell_type",
  fragment_files = FALSE,
  metacell_args = list(
    rna_reduction = "pca",
    rna_dims = 1:30,
    atac_reduction = "lsi",
    atac_dims = 2:30,
    gamma = 30,
    seed = 12345L,
    min_cells_per_stratum = 500,
    min_metacell_size = 10,
    min_metacells_per_stratum = 2L
  )
)

step3 <- rc_regcompass_step_meta_modules(
  grn = step1,
  metacells = step2,
  gem = gem,
  outdir = "RegCompass_steps/03_meta_modules"
)

step4 <- rc_regcompass_step_layer1(
  metacells = step2,
  meta_modules = step3,
  gem = gem,
  outdir = "RegCompass_steps/04_layer1",
  regulatory_alpha = 1,
  gpr_and_method = "min"
)

step5 <- rc_regcompass_step_layer2(
  layer1 = step4,
  meta_modules = step3,
  gem = gem,
  medium_scenarios = medium_scenarios,
  outdir = "RegCompass_steps/05_layer2",
  model_mode = "meta_module_gem",
  layer2_args = list(
    target_direction = "both",
    solver = "highs",
    model_params = list(
      completion_time_limit = 600,
      fastcore_epsilon = 1e-4,
      max_support_reactions = 2000,
      strict = TRUE
    )
  )
)

result <- rc_regcompass_step_results(
  grn = step1,
  metacells = step2,
  meta_modules = step3,
  layer1 = step4,
  layer2 = step5,
  gem = gem,
  outdir = "RegCompass_steps/06_results",
  species = "human"
)
```

# Inspect Stage 1 and Stage 2 provenance

```{r}
step1$grn_result$normalization_policy[c(
  "pando_motifs",
  "pando_regions",
  "pando_peak_cor",
  "pando_rsq"
)]

step2$pooled$cache_contract$analysis_args
step2$pooled$cache_contract$rna_reduction$embedding
step2$pooled$cache_contract$atac_reduction$embedding
```

# Inspect the Stage 3 evidence and catalogue

```{r}
step3$condition_modules$supported_metabolic_genes
step3$condition_modules$core_gene_reaction
step3$condition_modules$reaction_membership
step3$condition_modules$meta_module_summary$expansion_policy
step3$merged_modules$merged_core_reactions
step3$merged_modules$merged_reaction_membership
step3$merged_modules$is_gem
step3$merged_modules$fastcore_applied
```

The supported-gene table contains one row per condition, cell type, and significantly supported GEM target gene. It reports the number of significant edges, distinct TFs, distinct regions, minimum adjusted P value, maximum absolute coefficient, model R², and positive/negative edge counts.

The merged catalogue contains biological reactions only. It has no medium bounds and no FASTCORE support.

# Inspect Layer 1 GPR aggregation

```{r}
step4$capacity_params$and_method
step4$capacity_params$or_method
step4$evidence_formula
```

The default AND rule is `min`; `median` and `mean` are available for sensitivity analysis. The canonical OR rule remains additive.

# Inspect the Stage 5 union GEM

```{r}
step5$model_cache_summary[, c(
  "medium_scenario",
  "file",
  "file_checksum",
  "n_merged_biological_reactions",
  "n_global_fastcore_support_reactions",
  "completion_stage"
)]

union_gem <- readRDS(step5$model_cache_summary$file[[1]])

union_gem$is_union_gem
union_gem$union_gem_medium_scenario
union_gem$reaction_meta[
  union_gem$reaction_meta$global_fastcore_support %in% TRUE,
  , drop = FALSE
]
```

One union GEM is built per medium scenario and shared by all conditions and metacells in that scenario. The construction phase may be limited with `completion_time_limit`; scoring starts only after this cached model has been completed.

# Targeted second-pass scoring

```{r}
targeted <- rc_regcompass_step_target_union(
  layer1 = step4,
  meta_modules = step3,
  layer2 = step5,
  gem = gem,
  outdir = "RegCompass_targeted",
  core_reaction_ids = "MAR04324",
  layer2_args = list(
    target_direction = "both",
    solver = "highs"
  )
)

targeted$microcompass$params[c(
  "structural_model_reused_exactly",
  "fastcore_rerun",
  "model_rebuild",
  "scoring_time_limit"
)]
```

Selected genes or reaction IDs define mapping anchors. The second pass verifies each Stage 5 cache file checksum and medium identifier, reuses the exact final medium-specific union GEM, and neither rebuilds the model nor reruns FASTCORE. It is scoring-only and has no time-limit parameter.

# Restart policy

- Change Pando evidence thresholds, motifs, species, or regulatory regions: rerun Stage 1 onward.
- Change metacell cells, assays, reductions, dimensions, seed, gamma, or thresholds: rerun Stage 2 onward with `overwrite = TRUE`.
- Change subsystem annotations or GEM GPR rules: rerun Stage 3 onward.
- Change `gpr_and_method` or another RNA+ATAC transformation: rerun Stage 4 onward.
- Change medium or global FASTCORE construction settings: rerun Stage 5 onward.

# Condition comparisons

Within one medium, all conditions use the same union GEM and target-specific `vmax`. Condition differences therefore reflect the multiome penalty matrix. Different media may use different global FASTCORE support sets and must be interpreted as separate structural contexts.
