---
title: "RegCompassR 1.8.9 shared-background condition-sub-GRN workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RegCompassR 1.8.9 shared-background condition-sub-GRN workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

# Overview

RegCompassR 1.8.9 separates the shared regulatory hypothesis from condition-specific regulatory states.

```text
one validated Pando TF–peak–target structural universe per cell type
→ condition-aware TF×peak/target observability filter
→ one shared model edge universe across conditions
→ global backbone + symmetric condition deviations
→ leakage-resistant condition-stratified cross-validation
→ full-size condition-stratified bootstrap reproducibility
→ condition-specific metabolic target genes
→ complete-GPR condition cores
→ one ordered annotation expansion pass
→ one merged reaction catalogue
→ one medium-specific union GEM and one global FASTCORE completion
→ identical stoichiometric structure across conditions
→ condition/metacell-specific penalties and directional LP scores
→ compact final result
```

For edge `e`,

\[
\theta_{e,c}=\beta_e+\delta_{e,c},
\qquad
\sum_c\delta_{e,c}=0.
\]

Every condition uses the same Pando structural dictionary and the same observability-filtered model dictionary. Each condition contributes the same total regression loss weight. The canonical public workflow uses condition and cell type metadata only; it does not accept a biological-sample column.

# Prepare the GEM and medium

```{r}
library(RegCompassR)
library(BSgenome.Hsapiens.UCSC.hg38)

gem <- rc_prepare_gem(
  species = "human",
  version = "2.0.0",
  source = "bundled"
)

medium_scenarios <- rc_make_medium_scenarios(
  gem,
  scenario = "high_glucose",
  species = "human"
)
```

When `pfm` is omitted, RegCompass loads `data("motifs", package = "Pando")`. Human analyses use phastCons plus SCREEN ccRE regions; mouse analyses use the available phastCons region resource unless overridden.

# Complete workflow

```{r}
result <- rc_run_regcompass_one_shot(
  object = A,
  outdir = "RegCompass_result",
  genome = BSgenome.Hsapiens.UCSC.hg38,
  species = "human",
  gem = gem,
  condition_col = "Group",
  celltype_col = "cell_type",

  grn_mode = "multitask_shared_backbone",
  pando_args = list(
    min_cells = 100,
    pando_design_args = list(
      peak_to_gene_method = "Signac",
      upstream = 100000,
      downstream = 0,
      extend = 1000000,
      only_tss = FALSE,
      min_tf_detection = 0,
      min_peak_detection = 0,
      min_target_detection = 0,
      max_edges_per_target = Inf
    )
  ),
  multitask_args = list(
    alpha = 0.5,
    global_penalty_factor = 1,
    deviation_penalty_factor = 1,
    lambda_rule = "lambda.1se",
    nfolds = 5,
    n_bootstrap = 100,
    min_selection_frequency = 0.7,
    min_sign_stability = 0.8,
    min_bootstrap_success_fraction = 0.8,
    min_cv_rsq = 0,
    candidate_screen_threshold = 0,
    max_edges_per_target = Inf,
    min_detected_cells_per_condition = 10,
    min_detection_fraction_per_condition = 0.01,
    seed = 12345L
  ),

  fragment_files = FALSE,
  metacell_args = list(
    rna_reduction = "pca",
    rna_dims = 1:30,
    atac_reduction = "lsi",
    atac_dims = 2:30,
    gamma = 30,
    seed = 12345L,
    min_cells_per_stratum = 500,
    min_metacell_size = 10,
    min_metacells_per_stratum = 2L,
    overwrite = FALSE
  ),

  layer1_args = list(
    regulatory_alpha = 1,
    gpr_and_method = "min"
  ),

  medium_scenarios = medium_scenarios,
  model_mode = "meta_module_gem",
  layer2_args = list(
    target_direction = "both",
    solver = "highs",
    model_params = list(
      completion_time_limit = 600,
      fastcore_epsilon = 1e-4,
      max_support_reactions = 2000,
      strict = TRUE
    )
  ),
  upstream_workers = 6,
  layer2_workers = 30
)
```

# Structural and model candidate universes

`Pando::prepare_grn_design()` constructs structural candidates before fitting. An edge requires a target-linked regulatory region, a matched TF motif, an exact measured ATAC feature, and measured TF/target RNA features. Predictors are deduplicated by `(TF, ATAC feature, target)`, and the version-2 design fingerprint is validated before use.

The predictor is

\[
x_{e,u}=T_{t,u}A_{p,u}.
\]

Pando pooled detection thresholds remain zero so condition-restricted candidates are not removed prematurely. RegCompass defines

\[
m_c=\min\left(n_c,\max\left(10,\left\lceil0.01n_c\right\rceil\right)\right)
\]

and retains an edge for modelling only when, in at least one condition,

\[
\sum_{u\in c}I(T_{t,u}>0\land A_{p,u}>0)\ge m_c
\]

and

\[
\sum_{u\in c}I(Y_{g,u}>0)\ge m_c.
\]

TF and peak must therefore be observed together in the same cells. This detection filter does not use target correlation or fitted coefficients. The complete structural universe and the filtered model universe receive separate fingerprints.

Finite `max_edges_per_target` values are rejected in canonical mode because Pando candidate order is deterministic but not an evidence ranking.

# Condition model and predictive validation

Target and predictor values are centred within condition. Predictor scales are shared across conditions, while observation weights give every condition equal total loss.

The symmetric condition matrix is

\[
H=I_C-\frac1C\mathbf 1\mathbf 1^T.
\]

`alpha = 0.5` retains both lasso sparsity and ridge stabilization for correlated TF–peak predictors. Global and deviation coordinates use equal explicit penalty factors by default. A deviation factor above one is a sensitivity prior for a more conserved shared backbone, not a universal default.

For every cross-validation fold, condition centres and edge scales are estimated from training cells and applied to validation cells. Lambda uses the one-standard-error rule. A target model can produce active edges only when its out-of-fold

\[
R^2_{CV}>0.
\]

A positive `min_cv_rsq` adds a stronger predictive floor.

# Bootstrap reproducibility

For each bootstrap replicate, condition `c` is resampled with replacement at its original size `n_c`. The bootstrap target and predictors are re-centred within condition, scaled by the shared full-data scale, and fitted at the full-data selected lambda.

For successful bootstrap set \(\mathcal B_s\),

\[
\Pi_{e,c}=\frac1{|\mathcal B_s|}
\sum_{b\in\mathcal B_s}I(|\theta^{(b)}_{e,c}|>\varepsilon),
\]

\[
\rho_{e,c}=\left|
\frac{\sum_bI^{(b)}_{e,c}\operatorname{sign}(\theta^{(b)}_{e,c})}
{\sum_bI^{(b)}_{e,c}}
\right|,
\]

\[
\widetilde\theta_{e,c}
=\widehat\theta_{e,c}\Pi_{e,c}\rho_{e,c}.
\]

The canonical default is 100 bootstrap replicates. RegCompass reports selection-frequency Monte Carlo standard error and Wilson 95% intervals. `min_sign_stability = 0.8` requires at least 90% majority-sign agreement among selected fits because \(\rho=|2q-1|\).

Regularized coefficients do not receive classical adjusted p-values; `padj` remains `NA`. The full-size bootstrap quantifies cell-resampling reproducibility and does not claim formal half-sample stability-selection error control or biological-replicate significance.

# Condition genes and complete-GPR cores

A gene enters the condition target set when at least one active edge targets it. Positive and negative directions both establish membership.

For reaction `r` with GPR branches `B_{r,k}`:

\[
Core_{r,c}=1
\iff
\exists k:B_{r,k}\subseteq G_c.
\]

Stage 3 then performs exactly one ordered expansion:

```text
complete-GPR cores
→ core-reaction subsystems
→ direct KEGG/Reactome equivalents
→ direct master-Rhea equivalents
→ stop
```

# Layer 1 regulatory projection

Stage 1 uses RNA and ATAC to fit the GRN. Stage 4 reprojects only metacell ATAC. TF edges sharing one measured peak are signed-summed before projection. One target denominator is shared across conditions, preventing condition-wise rescaling of weak and strong sub-GRNs to equal magnitude.

A gene with no active condition edge has `R = 0`, and the log-odds update returns RNA-only support exactly:

\[
C^{MO}
=
\frac{C^{RNA}2^{\alpha R}}
{1-C^{RNA}+C^{RNA}2^{\alpha R}}.
\]

# Shared metabolic model

Condition core sets may differ, but Stage 3 merges their reaction catalogues before Stage 5. For each medium, Stage 5 constructs one union GEM and runs the only global FASTCORE completion. All conditions and metacells reuse identical reaction IDs, stoichiometric matrix, and bounds.

`completion_time_limit` limits FASTCORE union-GEM construction only. Directional scoring LPs have no time-limit parameter.

# Stepwise inspection

```{r}
step1 <- rc_regcompass_step_grn(
  object = A,
  gem = gem,
  outdir = "RegCompass_steps/01_grn",
  genome = BSgenome.Hsapiens.UCSC.hg38,
  species = "human",
  condition_col = "Group",
  celltype_col = "cell_type",
  grn_mode = "multitask_shared_backbone"
)

step2 <- rc_regcompass_step_metacells(
  object = A,
  outdir = "RegCompass_steps/02_metacells",
  condition_col = "Group",
  celltype_col = "cell_type",
  fragment_files = FALSE,
  metacell_args = list(
    rna_reduction = "pca",
    rna_dims = 1:30,
    atac_reduction = "lsi",
    atac_dims = 2:30,
    gamma = 30,
    seed = 12345L
  )
)

step3 <- rc_regcompass_step_meta_modules(
  grn = step1,
  metacells = step2,
  gem = gem,
  outdir = "RegCompass_steps/03_meta_modules"
)

step4 <- rc_regcompass_step_layer1(
  metacells = step2,
  meta_modules = step3,
  gem = gem,
  outdir = "RegCompass_steps/04_layer1",
  regulatory_alpha = 1,
  gpr_and_method = "min"
)

step5 <- rc_regcompass_step_layer2(
  layer1 = step4,
  meta_modules = step3,
  gem = gem,
  medium_scenarios = medium_scenarios,
  outdir = "RegCompass_steps/05_layer2",
  model_mode = "meta_module_gem",
  layer2_args = list(target_direction = "both", solver = "highs")
)

result <- rc_regcompass_step_results(
  grn = step1,
  metacells = step2,
  meta_modules = step3,
  layer1 = step4,
  layer2 = step5,
  gem = gem,
  outdir = "RegCompass_steps/06_results",
  species = "human"
)
```

Detailed Stage 1 outputs remain in `step1`:

```r
step1$grn_result$tf_peak_gene_candidates
step1$grn_result$tf_peak_gene_global
step1$grn_result$tf_peak_gene_condition_all
step1$grn_result$tf_peak_gene_significant
step1$grn_result$condition_target_genes
step1$grn_result$stability_diagnostics
step1$grn_result$group_status
```

The compact final result contains primary analysis tables:

```r
result$table_manifest
result$reaction_ranking
result$condition_contrast
result$active_regulatory_edges
result$condition_target_genes
result$core_reactions
result$reaction_catalog
result$reaction_evidence
result$microcompass$model_cache_summary
```

See `docs/grn-parameter-policy.md` for the canonical parameter derivations and `docs/multitask-shared-grn.md` for the complete mathematical and object contracts.
