---
title: "RegCompassR: from multiome input to directional scoring"
author: "RegCompass Developers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RegCompassR: from multiome input to directional scoring}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  purl = FALSE
)
```

## Workflow contract

RegCompassR exposes six supported functions. The explicit path prepares a
species GEM model with `rc_prepare_gem()`, defines exchange constraints
with `rc_make_medium_scenarios()`, and passes both to `rc_run_regcompass()`.
`rc_run_regcompass_one_shot()` performs those setup calls when `gem` or
`medium_scenarios` is omitted.

The integrated calculation follows this order:

```text
Seurat RNA + ATAC object plus optional fragment files
  -> condition x sample x cell-type strata
  -> fixed-resolution metacells within each stratum
  -> fragment re-aggregation when fragment files are supplied; otherwise object ATAC peak counts
  -> stratum-specific Pando regulatory evidence
  -> conservative GPR reaction evidence
  -> local FASTCORE-completed reaction meta-modules
  -> sample-balanced global Layer 1 and one shared GEM
  -> sample x cell-type directional minimum-discordance scoring
```

The primary score is an optimization penalty: lower values mean that the model
requires less disagreement with the observed evidence. It is not a measured
flux, a probability, or an independent observation for every metacell.

## 1. Validate the multiome input

`object` must inherit from `Seurat`, contain RNA and chromatin assays, and have
non-missing metadata columns for biological sample, condition, and cell type.
If matching fragment files are available, cell barcodes must agree across the
object and fragment files. Fragment files must be bgzip-compressed and
tabix-indexed. They may already be registered on the chromatin assay or be
supplied explicitly. If no matching fragment files are available, set
`fragment_files = FALSE`; RegCompass skips fragment aggregation and uses ATAC
peak raw counts already stored in the object.

```{r input-contract, purl=FALSE}
required_metadata <- c("sample_id", "condition", "cell_type")
stopifnot(inherits(object, "Seurat"))
stopifnot(all(required_metadata %in% colnames(object[[]])))
stopifnot(all(c("RNA", "ATAC") %in% names(object@assays)))
stopifnot(!anyNA(object[[]][, required_metadata]))

# A sample-aware manifest is the least ambiguous representation when fragments exist.
fragment_files <- data.frame(
  sample_id = c("donor_01", "donor_02"),
  assay = "ATAC",
  fragment_file = c(
    "fragments/donor_01.fragments.tsv.gz",
    "fragments/donor_02.fragments.tsv.gz"
  )
)
stopifnot(all(file.exists(fragment_files$fragment_file)))
stopifnot(all(file.exists(paste0(fragment_files$fragment_file, ".tbi"))))

with(object[[]], table(condition, sample_id, cell_type))
```

Every biological sample must contribute at least one retained stratum. Choose
`min_cells_per_stratum`, metacell resolution (`gamma`), and Pando's minimum
metacell count together; otherwise the global completion barrier will stop the
run rather than silently discard a sample.

## 2. Prepare and inspect the GEM

Pin a Human-GEM 2 release for reproducibility. The helper downloads, converts,
annotates, validates, and caches the model.

```{r gem, purl=FALSE}
library(RegCompassR)

human2_gem <- rc_prepare_human2_gem(
  version = "2.0.0",
  cache_dir = "model-cache"
)
# For mouse data, use rc_prepare_mouse_gem(version = "1.8.0").

dim(human2_gem$S)
head(human2_gem$gpr_table)
table(human2_gem$reaction_meta$role, useNA = "ifany")
human2_gem$model_info[c("source", "version", "commit", "checksum")]
```

Do not use `version = "latest"` for a reproducible analysis. If it is required
for exploration, it must be explicitly enabled with `allow_latest = TRUE`.

## 3. Define one shared medium

The default `compass_model_bounds` scenario preserves GEM exchange direction
and caps exchange bounds uniformly. It is a model-defined environment, not a
claim about measured metabolite concentrations.

Preset scenarios can also be returned together with a user-defined `"custom"` scenario in one medium table. Current optional named backgrounds are `normal_human_plasma`, `mouse_plasma`, `rpmi1640`, `dmem_high_glucose`, `low_glucose`, `high_glucose`, `low_lactate`, `high_lactate`, `low_glutamine`, `minimal`, and `permissive_all_exchange`. Removed compatibility aliases such as `human_plasma`, `hplm`, `blood_like`, `culture_like`, `tumor_low_glucose`, and `lactate_available` are intentionally unsupported; use canonical preset names or a custom medium table.

```{r medium, purl=FALSE}
medium <- rc_make_medium_scenarios(
  gem = human2_gem,
  scenario = "compass_model_bounds",
  exchange_limit = 1
)

stopifnot(all(medium$exchange_reaction_id %in% colnames(human2_gem$S)))
stopifnot(all(medium$lb <= medium$ub))
stopifnot(identical(unique(medium$condition), "all"))
head(medium[c("exchange_reaction_id", "lb", "ub", "available")])
```

Use a custom table only when bounds have an external justification. It must
include `medium_scenario_id`, `exchange_reaction_id`, `lb`, `ub`, and
`available`. Keep the medium shared across conditions unless the experimental
design independently establishes condition-specific exchange constraints.

## 4. Run the explicit workflow

The explicit entry point makes every major input visible. The PFM collection
and genome object shown below are typical human defaults; use objects matching
the organism and genome build of the ATAC assay.

```{r explicit-run, purl=FALSE}
library(Pando)
library(BSgenome.Hsapiens.UCSC.hg38)
data(motifs, package = "Pando")

result <- rc_run_regcompass(
  object = object,
  gem = human2_gem,
  outdir = "RegCompass_result",
  pfm = motifs,
  genome = BSgenome.Hsapiens.UCSC.hg38,
  fragment_files = FALSE,  # use object ATAC peak counts; pass fragment_files to re-aggregate
  sample_col = "sample_id",
  condition_col = "condition",
  celltype_col = "cell_type",
  rna_assay = "RNA",
  atac_assay = "ATAC",
  model_mode = "meta_module_gem",
  medium_scenarios = medium,
  metacell_args = list(
    gamma = 50,
    min_cells_per_stratum = 1000,
    min_metacell_size = 10
  ),
  pando_args = list(
    min_metacells = 20,
    pando_infer_args = list(
      method = "glm",
      tf_cor = 0.1,
      peak_cor = 0,
      adjust_method = "fdr"
    )
  ),
  layer1_args = list(
    local_fastcore = TRUE,
    sample_balance = TRUE,
    expression_batch_correction = "none"
  ),
  layer2_args = list(
    target_direction = "both",
    solver = "highs"
  ),
  inference_unit = "sample_celltype",
  upstream_workers = 4,
  layer2_workers = 4,
  parallel_backend = "snow"
)
```

The four `*_args` values are lists by contract. Integrated fields such as
`layer1`, `gem`, `unit`, and `medium_scenarios` are owned by the workflow and
must not be overridden through `layer2_args`. Use `inference_unit` on
`rc_run_regcompass()` to select the Layer 2 unit.

## 5. Equivalent one-shot entry point

For a first run, omit `gem` and `medium_scenarios`; the one-shot function
constructs them and delegates to the same integrated workflow. The default is
`species = "human"`; set `species = "mouse"` to automatically prepare
Mouse-GEM 1.8.0 and the mouse physiological medium. Arguments in `...` are
passed to `rc_run_regcompass()`.

```{r one-shot, purl=FALSE}
result <- rc_run_regcompass_one_shot(
  object = object,
  outdir = "RegCompass_result",
  pfm = motifs,
  genome = BSgenome.Hsapiens.UCSC.hg38,
  fragment_files = FALSE,  # or fragment_files to re-aggregate from fragments
  species = "human",
  gem_version = "2.0.0",
  medium_scenario = "compass_model_bounds",  # or "physiologic"/other presets
  sample_col = "sample_id",
  condition_col = "condition",
  celltype_col = "cell_type",
  inference_unit = "sample_celltype"
)
```

Set `fragment_files = FALSE` in the one-shot call when no matching fragment file
is available; pass the fragment manifest/path when fragments should be
re-aggregated. For mouse data, use a mouse genome and set `species = "mouse"`; the wrapper
then prepares Mouse-GEM 1.8.0 and the mouse physiological medium unless you
provide `gem` or `medium_scenarios`. Supply `gem = human2_gem` (or a Mouse-GEM
object) and `medium_scenarios = medium` to the one-shot function if those
objects were prepared separately. An explicit `medium_scenarios` table takes
precedence over `medium_scenario`.

## 6. Check the barrier and outputs

Do not begin differential analysis until the upstream barrier and sample
coverage have been inspected.

```{r inspect-result, purl=FALSE}
stopifnot(isTRUE(result$upstream_barrier$passed))
table(result$upstream_status$status, useNA = "ifany")

result$schema_version
result$params[c(
  "inference_unit",
  "medium_policy",
  "strict_biological_defaults",
  "penalty_unit"
)]

dim(result$layer1$C_abs)
head(result$grn_meta_modules$global_core_reactions)
dim(result$microcompass$penalty)
result$microcompass$primary_output_semantics
```

The returned object connects the phases as follows:

| Component | Produced by | Consumed by / interpretation |
|---|---|---|
| `upstream_status`, `upstream_barrier` | strict-stratum workers | verifies all required strata and samples reached the global phase |
| `layer1` | merged expression, GPR and Pando evidence | supplies absolute reaction evidence and diagnostics to Layer 2 |
| `grn_meta_modules` | local module construction and union | defines shared core and reaction membership for the meta-module GEM |
| `microcompass$penalty` | directional LP scoring | primary minimum evidence-discordance result; lower is more supported |
| `microcompass$relative_penalty_rank` | within-target ranking | display diagnostic, not the primary inferential quantity |

The run also writes the stratum manifest and status files, merged Layer 1,
global meta-modules, and the final result under `outdir`. Keep these artifacts
with the model version, medium table, parameters, and software environment.

## 7. Failure checkpoints

When a run stops, inspect the earliest applicable boundary:

1. **Input:** metadata completeness, assay names, fragment paths and indexes.
2. **Strata:** `00_strata/stratum_filter.tsv.gz` and
   `00_strata/stratum_workflow_status.tsv.gz`.
3. **Global barrier:** `00_strata/upstream_barrier.tsv.gz`; every sample must
   retain an analyzable stratum.
4. **Model:** reaction IDs in GPR, medium, membership, and target tables must
   match the stoichiometric matrix column names.
5. **Solver:** install the selected backend and record solver/time-limit
   settings; use `solver = "highs"` for the default open-source path.

For confirmatory comparisons, retain `inference_unit = "sample_celltype"` so
biological samples, rather than metacells from the same sample, are the
replicates.
