---
title: "RegCompassR 1.8.3 GRN-first workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RegCompassR 1.8.3 GRN-first workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

RegCompassR fits condition-by-cell-type Pando GRNs, builds condition-level SuperCell2 metacells, maps complete-GPR core reactions, constructs a FASTCORE-completed global union GEM, integrates RNA and ATAC evidence, and performs directional minimum-penalty LP scoring.

The canonical complete runner uses two automatic worker layers: `upstream_workers = 6L` and `layer2_workers = 30L`. It selects the operating-system backend, keeps every task inside an outer worker single-threaded, and releases each stage pool before the next unrelated stage.

## Installation

```{r}
install.packages(c("remotes", "highs", "BiocManager", "ggplot2"))
BiocManager::install("BiocParallel", ask = FALSE, update = FALSE)
remotes::install_github("1667857557/Pando_regcompass", upgrade = "never")
remotes::install_github("1667857557/Regcompass", upgrade = "never")
```

A local Pando archive is also accepted:

```{r}
install.packages("~/Pando_regcompass.tar.gz", repos = NULL, type = "source")
```

## Input, GEM, and medium

The Seurat object must contain paired RNA and ATAC counts for the same cells, complete condition/cell-type metadata, and an ATAC `ChromatinAssay`. Use `Pando::motifs`; do not pass `motif2tf` as `pfm`.

```{r}
library(RegCompassR)
library(Pando)
library(BSgenome.Hsapiens.UCSC.hg38)

data(motifs, package = "Pando")
condition_col <- "dataset"
celltype_col <- "epithelial_or_stem"

gem <- rc_prepare_gem(species = "human", version = "2.0.0")
medium_scenarios <- rc_make_medium_scenarios(
  gem = gem,
  scenario = "physiologic",
  species = "human"
)
```

On Linux, set numerical-library threads before launching R:

```bash
export OMP_NUM_THREADS=1
export OPENBLAS_NUM_THREADS=1
export MKL_NUM_THREADS=1
```

The package repeats this single-thread restriction during execution so package-managed workers inherit it.

## Complete run with automatic layered workers

```{r}
result <- rc_run_regcompass_one_shot(
  object = A,
  outdir = "RegCompass_result",
  pfm = motifs,
  genome = BSgenome.Hsapiens.UCSC.hg38,
  fragment_files = FALSE,
  species = "human",
  medium_scenarios = medium_scenarios,
  condition_col = condition_col,
  celltype_col = celltype_col,
  pando_args = list(
    min_cells = 100,
    pando_infer_args = list(
      method = "glm",
      tf_cor = 0.1,
      peak_cor = 0.01,
      adjust_method = "fdr",
      parallel = FALSE
    )
  ),
  metacell_args = list(
    gamma = 75,
    min_cells_per_stratum = 500,
    min_metacell_size = 10
  ),
  layer1_args = list(
    local_fastcore = TRUE,
    local_fastcore_args = list(
      solver = "highs",
      strict = TRUE,
      time_limit = 300
    )
  ),
  layer2_args = list(
    target_direction = "both",
    solver = "highs",
    time_limit = 60
  ),
  upstream_workers = 6L,
  layer2_workers = 30L,
  progress = TRUE
)
```

Do not supply `parallel_backend`, per-stage `BPPARAM` objects, or local-FASTCORE worker/backend fields to the complete runner. Windows automatically uses SOCK/SnowParam; Linux/macOS use MulticoreParam. Set both worker counts to `1L` for a fully serial run.

Every outer worker executes one single-thread analysis. Pando's inner `parallel = FALSE` is retained, numerical-library and nested R worker settings are fixed at one, and each package-managed worker pool is stopped and followed by `gc(full = TRUE)` after its stage.

## Audit saved stages

The complete run writes classed stage objects:

```{r}
step1 <- readRDS("RegCompass_result/01_single_cell_grn/step_grn.rds")
step2 <- readRDS("RegCompass_result/02_condition_metacells/step_metacells.rds")
step3 <- readRDS("RegCompass_result/03_meta_modules/step_meta_modules.rds")
step4 <- readRDS("RegCompass_result/04_layer1/step_layer1.rds")
step5 <- readRDS("RegCompass_result/05_layer2/step_layer2.rds")
result <- readRDS("RegCompass_result/06_results/regcompass_result.rds")
```

```{r}
stopifnot(
  inherits(step1, "regcompass_grn_step"),
  inherits(step2, "regcompass_metacell_step"),
  inherits(step3, "regcompass_meta_module_step"),
  inherits(step4, "regcompass_layer1_step"),
  inherits(step5, "regcompass_layer2_step"),
  all(step1$grn_result$sample_status$status == "ok"),
  !any(step2$pooled$metacell_meta$dominant_celltype_tied %in% TRUE),
  all(step3$group_coverage$coverage_complete),
  identical(step4$gem_fingerprint, step5$gem_fingerprint),
  identical(colnames(step4$reaction_expression), colnames(step5$penalty)),
  any(step5$evaluated),
  all(file.exists(step5$model_cache_summary$file)),
  identical(result$version, "1.8.3"),
  identical(result$params$upstream_workers, 6L),
  identical(result$params$layer2_workers, 30L),
  identical(result$params$internal_threads_per_task, 1L),
  identical(
    result$params$parallel_worker_lifecycle,
    "stage_scoped_create_start_stop_release_full_gc"
  )
)
```

Meta-modules contain complete-GPR cores, same-subsystem reactions, and reactions sharing KEGG, Reactome, or master-Rhea identifiers. FASTCORE adds separately labelled feasibility support. The raw minimum penalty is the primary result; the relative score is a within-target rank, not a probability.

## Direct database-linked non-core scoring

```{r}
expanded <- rc_regcompass_step_target_union(
  layer1 = step4,
  meta_modules = step3,
  layer2 = step5,
  gem = gem,
  outdir = "RegCompass_result/05b_expanded_targets",
  core_genes = c("GCLC", "GCLM", "GSS", "GSR", "G6PD", "PGD"),
  gene_match = "complete_gpr"
)

stopifnot(
  all(expanded$expanded_scoring_targets$score_target),
  all(expanded$expanded_reaction_catalog$expansion_type %in% c(
    "shared_kegg_reaction",
    "shared_reactome_reaction",
    "shared_master_rhea_reaction"
  )),
  identical(
    expanded$microcompass$params$target_scope,
    "direct_kegg_reactome_master_rhea_noncore_only"
  ),
  expanded$microcompass$params$structural_model_reused_exactly,
  all(expanded$microcompass$model_cache_summary$reused_without_rebuilding)
)
```

The selected core reactions are mapping anchors and are not scored again. Only non-core reactions directly sharing a KEGG, Reactome, or master-Rhea identifier with a selected core enter the second LP. Same-subsystem and transitive expansion are excluded. The exact original union-GEM files and bounds are reused.

For condition tests and plots, see `docs/condition-reaction-statistics.md`. For medium definitions, see `docs/medium-presets.md`. For restart and diagnostics, see `docs/tutorial-03-advanced-restart.md`.

Metacell-level condition comparisons describe separation within the analyzed dataset; they are not automatically biological-replicate-level treatment inference.
