---
title: "RegCompassR workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RegCompassR workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

# Overview

RegCompassR converts paired single-cell RNA+ATAC data into regulatory reaction
support and directional metabolic scores.

```text
>=2 condition levels -> condition_grn
0 or 1 condition level -> standard_pando using Pando::infer_grn()
```

Condition mode uses condition-full outer-heldout projection as the primary
regulatory route. Jointly estimable edges form the common-support component. A
non-estimable edge side remains projectable and contributes exactly zero.

Stage 2 uses:

```text
cell.graph.group = broad cell type
cell.split.condition = condition, or NULL when omitted
```

Each broad cell type has an independent graph; all its conditions share the
graph before condition-pure membership splitting. No sample or temporary
condition-by-cell-type stratum is used.

Mathematical definitions are centralized in
[Tutorial 3](../docs/tutorial-03-mathematical-model.md).

# Run

```{r}
library(RegCompassR)
library(BSgenome.Hsapiens.UCSC.hg38)

gem <- rc_prepare_gem(
  species = "human",
  version = "2.0.0",
  source = "bundled"
)

medium_scenarios <- rc_make_medium_scenarios(
  gem,
  scenario = "physiologic",
  species = "human"
)

result <- rc_run_regcompass(
  object = A,
  gem = gem,
  outdir = "RegCompass_result",
  genome = BSgenome.Hsapiens.UCSC.hg38,
  species = "human",
  condition_col = "Group",
  celltype_col = "cell_type",
  pando_args = list(
    min_cells = 100L,
    min_model_rsq = 0.1,
    pando_infer_args = list(
      candidate_screen = "motif_domain",
      condition_mix = 0.5,
      condition_weight = "equal",
      outer_nfolds = 5L,
      inner_nfolds = 5L,
      scale = TRUE
    )
  ),
  metacell_args = list(
    rna_reduction = "pca",
    rna_dims = 1:30,
    atac_reduction = "lsi",
    atac_dims = 2:30,
    gamma = 30L,
    seed = 12345L
  ),
  layer1_args = list(
    comparison_support = "auto",
    regulatory_alpha = 1,
    gpr_and_method = "min"
  ),
  medium_scenarios = medium_scenarios,
  model_mode = "meta_module_gem"
)
```

# Inspect

```{r}
result$analysis_mode
result$metacells$input_design
result$layer1$gene_projection_condition_full_oof
result$layer1$gene_projection_common_oof
result$microcompass$penalty_condition_full_oof
result$microcompass$penalty_common_oof
result$microcompass$penalty_condition_unique_increment
result$reaction_ranking
result$condition_contrast
```

`reaction_ranking` and `condition_contrast` use the condition-full penalty.
Common-support and RNA-only routes are decomposition/control outputs. All routes
reuse the exact same medium-specific model and `vmax`.

The canonical schema does not contain depth matching, common-depth restriction,
alpha sensitivity, zero-support sensitivity, or link-saturation propagation.

Metacells are within-dataset statistical units, not biological replicates.

# Optional targeted reaction remapping

After a completed `model_mode = "meta_module_gem"` run, directly database-linked
non-core reactions can be scored without rebuilding the model:

```{r}
targeted <- rc_regcompass_step_target_union(
  layer1 = result$layer1,
  meta_modules = result$meta_modules,
  layer2 = result$microcompass,
  gem = gem,
  outdir = "RegCompass_targeted",
  core_reaction_ids = c("MAR04381", "MAR04379"),
  layer2_args = list(target_direction = "both", solver = "highs")
)
```

The second pass reuses the exact Stage 5 union GEM and global FASTCORE
completion. In condition mode, `result$layer1$reaction_expression` is the
canonical alias of `reaction_expression_condition_full_oof`, so targeted targets
use the same primary evidence route as the original Stage 5 targets.

# Documentation

- [Tutorial 1](../docs/tutorial-01-quick-start.md)
- [Tutorial 2](../docs/tutorial-02-stepwise-audit.md)
- [Tutorial 3](../docs/tutorial-03-mathematical-model.md)
- [Tutorial 4: targeted reaction remapping](../docs/tutorial-04-targeted-reaction-remapping.md)
- [Tutorial 5: condition comparison](../docs/tutorial-05-condition-differential-analysis.md)
- [Stage contracts](../docs/stage-interface-contracts.md)
- [Public API](../docs/functions.md)
