---
title: "RegCompassR 1.8.0 GRN-first workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RegCompassR 1.8.0 GRN-first workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

RegCompassR first estimates condition-by-cell-type Pando GRNs from single cells. RNA is normalized once across all cells, while ATAC TF-IDF uses all conditions within each cell type as the shared reference. Pando then splits the normalized object by `condition_col × celltype_col`; the default peak-expression filter is `peak_cor = 0.01`.

After GRN inference, SuperCell2 builds `condition_col × celltype_col` metacells with `gamma = 75` by default. These metacells are used for Layer 1 reaction expression and Layer 2 scoring. Sample balancing and sample-based downsampling are not implemented.

```{r}
step1 <- rc_regcompass_step_grn(
  object, gem, "steps/01_grn", motifs,
  BSgenome.Hsapiens.UCSC.hg38,
  condition_col = "condition",
  celltype_col = "cell_type",
  pando_args = list(
    pando_infer_args = list(
      method = "glm", tf_cor = 0.1,
      peak_cor = 0.01, adjust_method = "fdr"
    )
  )
)

step2 <- rc_regcompass_step_metacells(
  object, "steps/02_metacells",
  condition_col = "condition",
  celltype_col = "cell_type",
  metacell_args = list(gamma = 75)
)

step3 <- rc_regcompass_step_meta_modules(
  step1, step2, gem, "steps/03_meta_modules"
)
step4 <- rc_regcompass_step_layer1(
  step2, step3, gem, "steps/04_layer1"
)
step5 <- rc_regcompass_step_layer2(
  step4, step3, gem, medium_scenarios,
  "steps/05_layer2"
)
result <- rc_regcompass_step_results(
  step1, step2, step3, step4, step5,
  gem, "steps/06_results"
)
```

Each condition-by-cell-type GRN defines metabolic-gene components and complete-GPR core reactions. Meta-modules then add reactions in the core-reaction subsystems and reactions sharing KEGG, Reactome, or master-Rhea identifiers. Local FASTCORE is the only feasibility-completion stage.
