---
title: "RegCompassR 1.8.1 GRN-first workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RegCompassR 1.8.1 GRN-first workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

RegCompassR first fits Pando GRNs on single cells and then constructs condition-only metacells. RNA is normalized once across all cells. ATAC TF-IDF is calculated within each cell type using all conditions as the shared reference. Pando is fitted separately for each `condition_col × celltype_col` group with `peak_cor = 0.01` by default. SuperCell2 then uses condition as its only varying stratum, with `gamma = 75` by default.

## Installation

```{r}
install.packages(c("remotes", "highs"))
remotes::install_github("1667857557/Pando_regcompass", upgrade = "never")
remotes::install_github("1667857557/Regcompass", upgrade = "never")
```

A downloaded Pando source archive is also valid:

```{r}
install.packages("~/Pando_regcompass.tar.gz", repos = NULL, type = "source")
```

RegCompass validates the Pando API and does not require GitHub remote metadata for local installations.

## Required object

The Seurat object must contain paired RNA and ATAC counts for the same cell IDs, an ATAC `ChromatinAssay`, complete condition and cell-type metadata, and peak coordinates matching the selected genome. Cell order may differ between assay matrices; RegCompass aligns normalized data by cell name.

`pfm` must be a motif PFM/PWM collection accepted by Pando and motifmatchr. Use `Pando::motifs`; do not pass the `motif2tf` annotation table.

```{r}
library(RegCompassR)
library(Pando)
library(BSgenome.Hsapiens.UCSC.hg38)

data(motifs, package = "Pando")

condition_col <- "dataset"
celltype_col <- "epithelial_or_stem"

stopifnot(
  inherits(A, "Seurat"),
  all(c("RNA", "ATAC") %in% names(A@assays)),
  inherits(A[["ATAC"]], "ChromatinAssay"),
  all(c(condition_col, celltype_col) %in% colnames(A@meta.data)),
  !anyNA(A@meta.data[[condition_col]]),
  !anyNA(A@meta.data[[celltype_col]])
)

gem <- rc_prepare_gem(species = "human", version = "2.0.0")
medium_scenarios <- rc_make_medium_scenarios(
  gem = gem,
  scenario = "high_glucose",
  species = "human"
)
```

## Step 1: single-cell GRNs

```{r}
step1 <- rc_regcompass_step_grn(
  object = A,
  gem = gem,
  outdir = "steps/01_grn",
  pfm = motifs,
  genome = BSgenome.Hsapiens.UCSC.hg38,
  condition_col = condition_col,
  celltype_col = celltype_col,
  pando_args = list(
    min_cells = 100,
    pando_infer_args = list(
      method = "glm",
      tf_cor = 0.1,
      peak_cor = 0.01,
      adjust_method = "fdr"
    )
  )
)

step1$grn_result$sample_status
head(step1$grn_result$tf_peak_gene_significant)
```

The stage writes `single_cell_grn.rds`, `step_grn.rds`, Pando group status, all/significant edge tables, and optional fitted Pando objects.

## Step 2: condition-only metacells

```{r}
step2 <- rc_regcompass_step_metacells(
  object = A,
  outdir = "steps/02_metacells",
  condition_col = condition_col,
  celltype_col = celltype_col,
  fragment_files = FALSE,
  metacell_args = list(
    gamma = 75,
    min_cells_per_stratum = 500,
    min_metacell_size = 10
  )
)

table(
  step2$pooled$metacell_meta[[condition_col]],
  step2$pooled$metacell_meta[[celltype_col]]
)
summary(step2$pooled$metacell_meta$dominant_celltype_fraction)
```

Cell type is assigned after metacell construction from member-cell composition. Exact dominant-cell-type ties are rejected. The stage writes the merged metacell Seurat object, final metadata, membership, and cell-type composition tables.

## Step 3: meta-modules

```{r}
step3 <- rc_regcompass_step_meta_modules(
  grn = step1,
  metacells = step2,
  gem = gem,
  outdir = "steps/03_meta_modules",
  layer1_args = list(
    local_fastcore = TRUE,
    local_fastcore_args = list(solver = "highs", strict = TRUE)
  )
)

step3$group_coverage
head(step3$condition_modules$core_gene_reaction)
step3$condition_modules$local_fastcore_summary
```

Each GRN defines metabolic-gene components and complete-GPR core reactions. Meta-modules add core-reaction subsystems and reactions sharing KEGG, Reactome, or master-Rhea identifiers. Local FASTCORE adds only feasibility-support reactions.

## Steps 4-6: reaction evidence and scoring

```{r}
step4 <- rc_regcompass_step_layer1(
  metacells = step2,
  meta_modules = step3,
  gem = gem,
  outdir = "steps/04_layer1"
)

step5 <- rc_regcompass_step_layer2(
  layer1 = step4,
  meta_modules = step3,
  gem = gem,
  medium_scenarios = medium_scenarios,
  outdir = "steps/05_layer2",
  layer2_args = list(
    target_direction = "both",
    solver = "highs"
  )
)

result <- rc_regcompass_step_results(
  grn = step1,
  metacells = step2,
  meta_modules = step3,
  layer1 = step4,
  layer2 = step5,
  gem = gem,
  outdir = "steps/06_results"
)

dim(step4$reaction_expression)
table(step5$feasible)
head(result$reaction_ranking)
```

The default HiGHS backend is checked before any medium-constrained model is constructed. A missing solver is reported as a solver installation problem rather than as GEM infeasibility.
