---
title: "RegCompassR workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RegCompassR workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

# Overview

RegCompassR converts paired single-cell RNA+ATAC data into condition-comparable
reaction support and directional metabolic scores.

```text
absolute condition-aware Pando GRNs
→ condition × broad-cell-type fixed-γ metacells
→ cell-type-specific RNA priors
→ structural-zero regulatory projection with RNA-only fallback
→ COMPASS GPR reaction expression and penalties
→ shared medium-specific metabolic model
→ directional scores and condition tests
```

Pando `ConditionGRNFit v5` outer-heldout condition projections are the primary
regulatory input. Conditions use absolute coefficients on one common coordinate;
no baseline-condition coefficient contrast is stored or consumed.

Mathematical definitions are centralized in
[Mathematical model](../docs/mathematical-model.md).

# Prepare GEM and medium

```{r}
library(RegCompassR)
library(BSgenome.Hsapiens.UCSC.hg38)

gem <- rc_prepare_gem(
  species = "human",
  version = "2.0.0",
  source = "bundled"
)

medium_scenarios <- rc_make_medium_scenarios(
  gem,
  scenario = "physiologic",
  species = "human"
)
```

Use `physiologic` for a species-specific baseline, a culture formulation for
in-vitro sensitivity analysis, or `custom` for experiment-specific exchange
bounds. See [Medium presets](../docs/medium-presets.md).

# One-shot workflow

```{r}
result <- rc_run_regcompass_one_shot(
  object = A,
  outdir = "RegCompass_result",
  genome = BSgenome.Hsapiens.UCSC.hg38,
  species = "human",
  gem = gem,
  condition_col = "Group",
  celltype_col = "cell_type",
  pando_args = list(
    min_cells = 100L,
    min_abs_estimate = 0,
    min_model_rsq = 0.1,
    pando_infer_args = list(
      candidate_screen = "motif_domain",
      condition_mix = 0.5,
      condition_weight = "equal",
      outer_nfolds = 5L,
      inner_nfolds = 5L,
      lambda_selection = "lambda.1se",
      scale = TRUE
    )
  ),
  fragment_files = FALSE,
  metacell_args = list(
    rna_reduction = "pca",
    rna_dims = 1:30,
    atac_reduction = "lsi",
    atac_dims = 2:30,
    gamma = 30L,
    seed = 12345L,
    min_cells_per_stratum = 500L,
    min_metacell_size = 10L,
    min_metacells_per_stratum = 2L
  ),
  layer1_args = list(
    projection_component = "condition",
    comparison_support = "auto",
    regulatory_alpha = 1,
    gpr_and_method = "min"
  ),
  medium_scenarios = medium_scenarios,
  model_mode = "meta_module_gem",
  layer2_args = list(
    target_direction = "both",
    solver = "highs",
    model_params = list(
      completion_time_limit = 600,
      fastcore_epsilon = 1e-4,
      max_support_reactions = 2000,
      strict = TRUE
    )
  ),
  upstream_workers = 6L,
  layer2_workers = 30L
)
```

The one-shot runner controls Stage 1 parallelism. Do not put `parallel` or
`BPPARAM` inside `pando_infer_args`.

Every condition × broad-cell-type stratum uses the same `gamma = 30L`. RNA/ATAC
depth does not alter gamma, and top-1% depth cells are QC annotations only.

The RNA Gamma–Poisson prior is estimated by broad cell type. A non-estimable
Pando edge contributes a structural zero in the primary projection path. A
non-finite target-level modifier falls back exactly to RNA-only support.
`regulatory_alpha = 1` is fixed.

COMPASS GPR semantics are used: the selected AND rule is applied within
complexes, OR isozyme branches are summed while unavailable branches are
ignored, and missing final reaction expression receives the maximum
expression-linked penalty.

# Main outputs

```{r}
result$grn$condition_fit_status
result$grn$condition_grn_fits
result$grn$tf_peak_gene_condition
result$condition_grn_meta_modules$supported_metabolic_genes
result$condition_grn_meta_modules$core_gene_reaction
result$layer1$projection_structural_zero
result$layer1$regulatory_fallback
result$microcompass$model_cache_summary
result$reaction_ranking
result$condition_contrast
result$reaction_comparison_by_metacell
```

`result$condition_contrast` is a downstream metabolic comparison, not a Pando
coefficient contrast.

# Stepwise workflow

```{r}
library(BiocParallel)

upstream_bp <- if (.Platform$OS.type == "windows") {
  SnowParam(workers = 6L, type = "SOCK")
} else {
  MulticoreParam(workers = 6L)
}

step1 <- rc_regcompass_step_grn(
  object = A,
  gem = gem,
  outdir = "RegCompass_steps/01_grn",
  genome = BSgenome.Hsapiens.UCSC.hg38,
  species = "human",
  condition_col = "Group",
  celltype_col = "cell_type",
  pando_args = list(
    min_cells = 100L,
    min_model_rsq = 0.1,
    pando_infer_args = list(
      candidate_screen = "motif_domain",
      condition_mix = 0.5,
      condition_weight = "equal"
    )
  ),
  parallel = TRUE,
  BPPARAM = upstream_bp
)

step2 <- rc_regcompass_step_metacells(
  object = A,
  outdir = "RegCompass_steps/02_metacells",
  condition_col = "Group",
  celltype_col = "cell_type",
  fragment_files = FALSE,
  metacell_args = list(
    rna_reduction = "pca",
    rna_dims = 1:30,
    atac_reduction = "lsi",
    atac_dims = 2:30,
    gamma = 30L,
    seed = 12345L,
    min_cells_per_stratum = 500L,
    min_metacell_size = 10L,
    min_metacells_per_stratum = 2L
  )
)

step3 <- rc_regcompass_step_meta_modules(
  grn = step1,
  metacells = step2,
  gem = gem,
  outdir = "RegCompass_steps/03_meta_modules"
)

step4 <- rc_regcompass_step_layer1(
  grn = step1,
  metacells = step2,
  meta_modules = step3,
  gem = gem,
  outdir = "RegCompass_steps/04_layer1",
  projection_component = "condition",
  comparison_support = "auto",
  regulatory_alpha = 1,
  gpr_and_method = "min",
  parallel = TRUE,
  BPPARAM = upstream_bp
)

step5 <- rc_regcompass_step_layer2(
  layer1 = step4,
  meta_modules = step3,
  gem = gem,
  medium_scenarios = medium_scenarios,
  outdir = "RegCompass_steps/05_layer2",
  model_mode = "meta_module_gem",
  layer2_args = list(
    target_direction = "both",
    solver = "highs",
    model_params = list(
      completion_time_limit = 600,
      fastcore_epsilon = 1e-4,
      max_support_reactions = 2000,
      strict = TRUE
    )
  )
)

result <- rc_regcompass_step_results(
  grn = step1,
  metacells = step2,
  meta_modules = step3,
  layer1 = step4,
  layer2 = step5,
  gem = gem,
  outdir = "RegCompass_steps/06_results",
  species = "human"
)
```

# Condition comparison

```{r}
comparison <- rc_test_condition_reactions(
  result,
  condition_col = "Group",
  reaction_ids = "MAR04324",
  target_directions = "forward",
  medium_scenarios = "physiologic",
  comparisons = list(c("Control", "Drug"))
)

comparison$pairwise
comparison$omnibus
```

Compare the same reaction, direction, medium, and broad cell type. Metacells are
the statistical units; P values describe within-dataset separation and are not
donor-level inference.

# Mouse input

Mouse runs require a regulatory-region object matching the ATAC coordinates and
the motif-scanning genome:

```{r}
library(BSgenome.Mmusculus.UCSC.mm10)

mouse_regions <- readRDS("mm10_regulatory_regions.rds")

mouse_result <- rc_run_regcompass_one_shot(
  object = A_mouse,
  outdir = "RegCompass_mouse",
  genome = BSgenome.Mmusculus.UCSC.mm10,
  species = "mouse",
  condition_col = "Group",
  celltype_col = "cell_type",
  pando_args = list(
    pando_initiate_args = list(regions = mouse_regions),
    pando_infer_args = list(
      candidate_screen = "motif_domain",
      condition_mix = 0.5,
      condition_weight = "equal"
    )
  ),
  metacell_args = list(
    rna_reduction = "pca",
    atac_reduction = "lsi",
    gamma = 30L,
    seed = 12345L
  ),
  layer1_args = list(
    projection_component = "condition",
    comparison_support = "auto",
    regulatory_alpha = 1,
    gpr_and_method = "min"
  )
)
```

# Further documentation

- [Workflow](../docs/workflow.md)
- [Pando condition contract](../docs/condition-comparable-grn.md)
- [Restart and sensitivity](../docs/tutorial-03-advanced-restart.md)
- [Medium presets](../docs/medium-presets.md)
- [Public API](../docs/functions.md)
