---
title: "RegCompassR 1.8.8 shared-background condition-sub-GRN workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RegCompassR 1.8.8 shared-background condition-sub-GRN workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

# Overview

RegCompassR 1.8.8 separates the shared regulatory hypothesis from condition-specific regulatory states.

```text
one Pando structural TF–peak–target universe per cell type
→ global backbone + symmetric condition deviations
→ stability-selected condition sub-GRNs
→ condition-specific metabolic target genes
→ complete-GPR condition cores
→ one ordered annotation expansion pass
→ one merged reaction catalogue
→ one medium-specific union GEM and one global FASTCORE completion
→ identical stoichiometric structure across conditions
→ condition/metacell-specific penalties and directional LP scores
```

The condition coefficient for edge `e` is

\[
\theta_{e,c}=\beta_e+\delta_{e,c},
\qquad
\sum_c\delta_{e,c}=0.
\]

Every condition uses the same Pando candidate dictionary. Observation weights are inversely proportional to the number of cells in each condition, so every condition contributes the same total loss weight.

# Prepare the GEM and medium

```{r}
library(RegCompassR)
library(BSgenome.Hsapiens.UCSC.hg38)

gem <- rc_prepare_gem(
  species = "human",
  version = "2.0.0",
  source = "bundled"
)

medium_scenarios <- rc_make_medium_scenarios(
  gem,
  scenario = "high_glucose",
  species = "human"
)
```

When `pfm` is omitted, RegCompass loads:

```r
data("motifs", package = "Pando")
pfm <- motifs
```

Human analyses use the union of `phastConsElements20Mammals.UCSC.hg38` and `SCREEN.ccRE.UCSC.hg38` unless an explicit region object is supplied. Mouse analyses use the bundled phastCons object by default.

# Complete workflow

```{r}
result <- rc_run_regcompass_one_shot(
  object = A,
  outdir = "RegCompass_result",
  genome = BSgenome.Hsapiens.UCSC.hg38,
  species = "human",
  gem = gem,
  condition_col = "Group",
  celltype_col = "cell_type",
  sample_col = "sample_id",

  grn_mode = "multitask_shared_backbone",
  pando_args = list(
    min_cells = 100,
    pando_design_args = list(
      peak_to_gene_method = "Signac",
      min_tf_detection = 0.01,
      min_peak_detection = 0.01,
      min_target_detection = 0.01
    )
  ),
  multitask_args = list(
    alpha = 0.5,
    global_penalty_factor = 1,
    deviation_penalty_factor = 2,
    lambda_rule = "lambda.1se",
    nfolds = 5,
    n_stability = 50,
    stability_fraction = 0.8,
    min_selection_frequency = 0.7,
    min_sign_stability = 0.8,
    candidate_screen_threshold = 0,
    max_edges_per_target = Inf,
    seed = 12345L
  ),

  fragment_files = FALSE,
  metacell_args = list(
    rna_reduction = "pca",
    rna_dims = 1:30,
    atac_reduction = "lsi",
    atac_dims = 2:30,
    gamma = 30,
    seed = 12345L,
    min_cells_per_stratum = 500,
    min_metacell_size = 10,
    min_metacells_per_stratum = 2L,
    overwrite = FALSE
  ),

  layer1_args = list(
    regulatory_alpha = 1,
    gpr_and_method = "min"
  ),

  medium_scenarios = medium_scenarios,
  model_mode = "meta_module_gem",
  layer2_args = list(
    target_direction = "both",
    solver = "highs",
    model_params = list(
      completion_time_limit = 600,
      fastcore_epsilon = 1e-4,
      max_support_reactions = 2000,
      strict = TRUE
    )
  ),
  upstream_workers = 6,
  layer2_workers = 30
)
```

# Shared candidate background

`Pando::prepare_grn_design()` constructs structural candidates before fitting. An edge requires a target-linked Pando regulatory region, a matched TF motif, a measured ATAC feature, and measured TF/target RNA features. Exact predictors are deduplicated by `(TF, ATAC feature, target)`.

The Pando-style predictor is

\[
x_{e,u}=T_{t,u}A_{p,u}.
\]

The default `candidate_screen_threshold = 0` and `max_edges_per_target = Inf` preserve the full structural universe. A positive screen threshold is optional and uses the maximum within-condition absolute correlation, but every condition still receives the same retained dictionary.

# Condition model and stability

Target and predictor values are centred within condition, or within `condition × sample` when `sample_col` is supplied. Shared scaling is calculated across all conditions of the cell type.

The elastic-net fit uses the symmetric centring matrix

\[
H=I_C-\frac1C\mathbf 1\mathbf 1^T.
\]

Because the `C` symmetric deviation columns are rank deficient, `alpha` must be below one. The ridge component makes the solution unique and invariant to the choice of reference condition.

Repeated stratified subsampling gives

\[
\widetilde\theta_{e,c}
=
\widehat\theta_{e,c}
\Pi_{e,c}
\rho_{e,c},
\]

where `Pi` is the selection frequency and `rho` is conditional sign stability. The multitask workflow does not attach a classical adjusted p-value to regularised coefficients; `padj` remains `NA`.

# Condition genes and complete-GPR cores

A gene enters the condition target set when at least one active stable edge targets it. Positive and negative regulatory directions both establish membership.

For reaction `r` with GPR branches `B_rk`:

\[
Core_{r,c}=1
\iff
\exists k:B_{r,k}\subseteq G_c.
\]

Stage 3 then performs exactly one ordered expansion:

```text
complete-GPR cores
→ core-reaction subsystems
→ direct KEGG/Reactome equivalents
→ direct master-Rhea equivalents
→ stop
```

# Layer 1 regulatory projection

Stage 1 uses RNA and ATAC to fit the GRN. Stage 4 reprojects only metacell ATAC. TF edges sharing the same measured peak are signed-summed before projection. One target-specific denominator is shared across conditions, avoiding condition-wise rescaling of weak and strong sub-GRNs to the same magnitude.

A gene with no active condition edge has modifier `R = 0`. The existing log-odds update then returns RNA-only support exactly:

\[
C^{MO}
=
\frac{C^{RNA}2^{\alpha R}}
{1-C^{RNA}+C^{RNA}2^{\alpha R}}.
\]

# Shared metabolic model

Condition-specific core sets may differ, but Stage 3 merges their reaction catalogues before Stage 5. For each medium, Stage 5 constructs one union GEM and runs the only global FASTCORE completion. All conditions and metacells reuse identical reaction IDs, stoichiometric matrix and bounds.

`completion_time_limit` limits FASTCORE union-GEM construction only. Directional scoring LPs do not have a time-limit parameter.

# Stepwise inspection

```{r}
step1 <- rc_regcompass_step_grn(
  object = A,
  gem = gem,
  outdir = "RegCompass_steps/01_grn",
  genome = BSgenome.Hsapiens.UCSC.hg38,
  species = "human",
  sample_col = "sample_id",
  condition_col = "Group",
  celltype_col = "cell_type",
  grn_mode = "multitask_shared_backbone",
  pando_args = list(
    min_cells = 100,
    pando_design_args = list(
      peak_to_gene_method = "Signac"
    )
  ),
  multitask_args = list(
    alpha = 0.5,
    n_stability = 50,
    max_edges_per_target = Inf,
    seed = 12345L
  )
)

step2 <- rc_regcompass_step_metacells(
  object = A,
  outdir = "RegCompass_steps/02_metacells",
  sample_col = "sample_id",
  condition_col = "Group",
  celltype_col = "cell_type",
  fragment_files = FALSE,
  metacell_args = list(
    rna_reduction = "pca",
    rna_dims = 1:30,
    atac_reduction = "lsi",
    atac_dims = 2:30,
    gamma = 30,
    seed = 12345L
  )
)

step3 <- rc_regcompass_step_meta_modules(
  grn = step1,
  metacells = step2,
  gem = gem,
  outdir = "RegCompass_steps/03_meta_modules"
)

step4 <- rc_regcompass_step_layer1(
  metacells = step2,
  meta_modules = step3,
  gem = gem,
  outdir = "RegCompass_steps/04_layer1",
  regulatory_alpha = 1,
  gpr_and_method = "min"
)

step5 <- rc_regcompass_step_layer2(
  layer1 = step4,
  meta_modules = step3,
  gem = gem,
  medium_scenarios = medium_scenarios,
  outdir = "RegCompass_steps/05_layer2",
  model_mode = "meta_module_gem",
  layer2_args = list(target_direction = "both", solver = "highs")
)

result <- rc_regcompass_step_results(
  grn = step1,
  metacells = step2,
  meta_modules = step3,
  layer1 = step4,
  layer2 = step5,
  gem = gem,
  outdir = "RegCompass_steps/06_results",
  species = "human"
)
```

The principal Stage 1 tables are:

```r
result$grn$tf_peak_gene_candidates
result$grn$tf_peak_gene_global
result$grn$tf_peak_gene_condition_all
result$grn$tf_peak_gene_significant
result$grn$condition_target_genes
```

The condition and merged metabolic outputs are:

```r
result$condition_grn_meta_modules$supported_metabolic_genes
result$condition_grn_meta_modules$core_gene_reaction
result$merged_grn_meta_modules$merged_core_reactions
result$merged_grn_meta_modules$merged_reaction_membership
result$microcompass$model_cache_summary
```

See `docs/multitask-shared-grn.md` for the complete equations and object contracts.
