---
title: "RegCompassR 1.8.4 significant-Pando-target workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RegCompassR 1.8.4 significant-Pando-target workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

# Overview

RegCompassR converts paired single-cell RNA+ATAC evidence into reaction-level support and directional COMPASS-like scores.

```text
condition × cell type Pando TF–peak–Human-GEM-gene models
→ significant metabolic target-gene sets
→ complete-GPR core reactions
→ annotation-defined biological meta-modules
→ merged meta-module catalogue
→ integrated RNA+ATAC reaction support
→ medium-specific union GEM
→ single global FASTCORE completion
→ directional LP scoring
```

Stage 3 does not project targets through shared TFs, calculate GRN connected components, run FASTCORE, or create a GEM. Only the medium-constrained Stage 5 model is called a union GEM.

# Prepare the GEM and medium

```{r}
library(RegCompassR)
library(BSgenome.Hsapiens.UCSC.hg38)

data(motif2tf, package = "Pando")

gem <- rc_prepare_gem(
  species = "human",
  version = "2.0.0",
  source = "bundled"
)

medium_scenarios <- rc_make_medium_scenarios(
  gem,
  scenario = "high_glucose",
  species = "human"
)
```

For human analyses, RegCompass loads `phastConsElements20Mammals.UCSC.hg38` and `SCREEN.ccRE.UCSC.hg38` from Pando, takes their union, and passes it to `Pando::initiate_grn(regions = ...)`. Supply `pando_args$pando_initiate_args$regions` to override this default. Non-human analyses require species-appropriate regions explicitly.

# Complete workflow

```{r}
result <- rc_run_regcompass_one_shot(
  object = A,
  outdir = "RegCompass_result",
  pfm = motif2tf,
  genome = BSgenome.Hsapiens.UCSC.hg38,
  fragment_files = FALSE,
  gem = gem,
  species = "human",
  medium_scenarios = medium_scenarios,
  sample_col = NULL,
  condition_col = "Group",
  celltype_col = "cell_type",
  model_mode = "meta_module_gem",
  metacell_args = list(
    gamma = 30,
    min_cells_per_stratum = 500,
    min_metacell_size = 10
  ),
  pando_args = list(
    min_cells = 100,
    padj_threshold = 0.05,
    min_abs_estimate = 0,
    min_model_rsq = 0.1,
    pando_infer_args = list(
      method = "glm",
      tf_cor = 0.1,
      peak_cor = 0.01,
      adjust_method = "fdr",
      parallel = FALSE
    )
  ),
  meta_module_args = list(
    expansion_mode = "ordered_once"
  ),
  layer1_args = list(
    regulatory_alpha = 1,
    tau = 0.20
  ),
  layer2_args = list(
    target_direction = "both",
    solver = "highs",
    model_params = list(
      completion_time_limit = 600,
      fastcore_epsilon = 1e-4,
      max_support_reactions = 2000,
      strict = TRUE
    )
  ),
  upstream_workers = 6,
  layer2_workers = 30
)
```

The Stage 1 adjusted-P-value, effect-size, and model-R² filters define which Human-GEM target genes enter Stage 3. Both positive and negative Pando coefficients count as regulatory evidence. FASTCORE is configured only at Stage 5 through `layer2_args$model_params`. `completion_time_limit` limits union-GEM construction only; the later scoring LPs have no time-limit parameter.

# Stepwise workflow

```{r}
step1 <- rc_regcompass_step_grn(
  object = A,
  gem = gem,
  outdir = "RegCompass_steps/01_grn",
  pfm = motif2tf,
  genome = BSgenome.Hsapiens.UCSC.hg38,
  condition_col = "Group",
  celltype_col = "cell_type",
  pando_args = list(
    min_cells = 100,
    padj_threshold = 0.05,
    min_model_rsq = 0.1
  )
)

step2 <- rc_regcompass_step_metacells(
  object = A,
  outdir = "RegCompass_steps/02_metacells",
  sample_col = NULL,
  condition_col = "Group",
  celltype_col = "cell_type",
  fragment_files = FALSE,
  metacell_args = list(gamma = 30)
)

step3 <- rc_regcompass_step_meta_modules(
  grn = step1,
  metacells = step2,
  gem = gem,
  outdir = "RegCompass_steps/03_meta_modules",
  meta_module_args = list(expansion_mode = "ordered_once")
)

step4 <- rc_regcompass_step_layer1(
  metacells = step2,
  meta_modules = step3,
  gem = gem,
  outdir = "RegCompass_steps/04_layer1",
  regulatory_alpha = 1,
  tau = 0.20
)

step5 <- rc_regcompass_step_layer2(
  layer1 = step4,
  meta_modules = step3,
  gem = gem,
  medium_scenarios = medium_scenarios,
  outdir = "RegCompass_steps/05_layer2",
  model_mode = "meta_module_gem",
  layer2_args = list(
    target_direction = "both",
    solver = "highs",
    model_params = list(
      completion_time_limit = 600,
      fastcore_epsilon = 1e-4,
      max_support_reactions = 2000,
      strict = TRUE
    )
  )
)

result <- rc_regcompass_step_results(
  grn = step1,
  metacells = step2,
  meta_modules = step3,
  layer1 = step4,
  layer2 = step5,
  gem = gem,
  outdir = "RegCompass_steps/06_results",
  species = "human"
)
```

# Inspect the Stage 3 evidence and catalogue

```{r}
step3$condition_modules$supported_metabolic_genes
step3$condition_modules$core_gene_reaction
step3$condition_modules$reaction_membership
step3$merged_modules$merged_core_reactions
step3$merged_modules$merged_reaction_membership
step3$merged_modules$is_gem
step3$merged_modules$fastcore_applied
```

The supported-gene table contains one row per condition, cell type, and significantly supported Human-GEM target gene. It reports the number of significant edges, distinct TFs, distinct regions, minimum adjusted P value, maximum absolute coefficient, model R², and positive/negative edge counts.

The merged catalogue contains biological reactions only. It has no medium bounds and no FASTCORE support.

# Inspect the Stage 5 union GEM

```{r}
step5$model_cache_summary[, c(
  "medium_scenario",
  "file",
  "file_checksum",
  "n_merged_biological_reactions",
  "n_global_fastcore_support_reactions",
  "completion_stage"
)]

union_gem <- readRDS(step5$model_cache_summary$file[[1]])

union_gem$is_union_gem
union_gem$union_gem_medium_scenario
union_gem$reaction_meta[
  union_gem$reaction_meta$global_fastcore_support %in% TRUE,
  , drop = FALSE
]
```

One union GEM is built per medium scenario and shared by all conditions and metacells in that scenario. The construction phase may be limited with `completion_time_limit`; scoring starts only after this cached model has been completed.

# Targeted second-pass scoring

```{r}
targeted <- rc_regcompass_step_target_union(
  layer1 = step4,
  meta_modules = step3,
  layer2 = step5,
  gem = gem,
  outdir = "RegCompass_targeted",
  core_reaction_ids = "MAR04324",
  layer2_args = list(
    target_direction = "both",
    solver = "highs"
  )
)

targeted$microcompass$params[c(
  "structural_model_reused_exactly",
  "fastcore_rerun",
  "model_rebuild",
  "scoring_time_limit"
)]
```

Selected genes or reaction IDs define mapping anchors. The second pass verifies each Stage 5 cache file checksum and medium identifier, reuses the exact final medium-specific union GEM, and neither rebuilds the model nor reruns FASTCORE. It is scoring-only and has no time-limit parameter.

# Restart policy

- Change Pando evidence thresholds or regulatory regions: rerun Stage 1 onward.
- Change subsystem or reaction-equivalence expansion: rerun Stage 3 onward.
- Change the RNA+ATAC transformation: rerun Stage 4 onward.
- Change medium or global FASTCORE construction settings: rerun Stage 5 onward.

# Condition comparisons

Within one medium, all conditions use the same union GEM and target-specific `vmax`. Condition differences therefore reflect the multiome penalty matrix. Different media may use different global FASTCORE support sets and must be interpreted as separate structural contexts.
