---
title: "RegCompassR 1.8.10 shared-background condition-sub-GRN workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RegCompassR 1.8.10 shared-background condition-sub-GRN workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

# Overview

RegCompassR 1.8.10 separates a shared structural regulatory hypothesis from condition-specific sparse coefficients and records the biological unit used for bootstrap stability.

```text
one validated GREAT-domain Pando TF–peak–target universe per cell type
→ condition-aware TF×peak/target observability filter
→ one shared model edge universe across conditions
→ direct condition-specific theta elastic net
→ leakage-resistant condition-stratified cell-level cross-validation
→ condition-stratified sample/donor cluster bootstrap when sample_col is valid
→ explicit warning and condition-stratified cell-bootstrap fallback otherwise
→ condition-specific metabolic target genes
→ complete-GPR condition cores
→ one merged reaction catalogue
→ one medium-specific union GEM and one global FASTCORE completion
→ identical stoichiometric structure across conditions
→ condition/metacell-specific penalties and directional LP scores
→ compact final result
```

For edge \(e\), RegCompass estimates \(\theta_{e,c}\). The exported backbone and deviation are derived:

\[
\beta_e=\frac1C\sum_c\theta_{e,c},
\qquad
\delta_{e,c}=\theta_{e,c}-\beta_e,
\qquad
\sum_c\delta_{e,c}=0.
\]

Every condition uses the same Pando structural and observability-filtered dictionaries. The optional biological sample column affects bootstrap resampling only. Stage 2 remains condition-only.

# Prepare the GEM and medium

```{r}
library(RegCompassR)
library(BSgenome.Hsapiens.UCSC.hg38)

gem <- rc_prepare_gem(
  species = "human",
  version = "2.0.0",
  source = "bundled"
)

medium_scenarios <- rc_make_medium_scenarios(
  gem,
  scenario = "high_glucose",
  species = "human"
)
```

# Complete workflow

```{r}
result <- rc_run_regcompass_one_shot(
  object = A,
  outdir = "RegCompass_result",
  genome = BSgenome.Hsapiens.UCSC.hg38,
  species = "human",
  gem = gem,
  condition_col = "Group",
  celltype_col = "cell_type",
  sample_col = "sample_id",
  grn_mode = "multitask_shared_backbone",
  pando_args = list(
    min_cells = 100,
    pando_design_args = list(
      peak_to_gene_method = "GREAT",
      upstream = 100000,
      downstream = 0,
      extend = 1000000,
      only_tss = FALSE,
      min_tf_detection = 0,
      min_peak_detection = 0,
      min_target_detection = 0,
      max_edges_per_target = Inf
    )
  ),
  multitask_args = list(
    alpha = 0.5,
    global_penalty_factor = 1,
    deviation_penalty_factor = 1,
    lambda_rule = "lambda.1se",
    nfolds = 5,
    n_bootstrap = 100,
    min_selection_frequency = 0.7,
    min_sign_stability = 0.8,
    min_bootstrap_success_fraction = 0.8,
    min_cv_rsq = 0,
    candidate_screen_threshold = 0,
    max_edges_per_target = Inf,
    min_detected_cells_per_condition = 10,
    min_detection_fraction_per_condition = 0.01,
    seed = 12345L
  ),
  fragment_files = FALSE,
  metacell_args = list(
    rna_reduction = "pca",
    rna_dims = 1:30,
    atac_reduction = "lsi",
    atac_dims = 2:30,
    gamma = 30,
    seed = 12345L,
    min_cells_per_stratum = 500,
    min_metacell_size = 10,
    min_metacells_per_stratum = 2L,
    overwrite = FALSE
  ),
  layer1_args = list(
    regulatory_alpha = 1,
    gpr_and_method = "min"
  ),
  medium_scenarios = medium_scenarios,
  model_mode = "meta_module_gem",
  layer2_args = list(
    target_direction = "both",
    solver = "highs",
    model_params = list(
      completion_time_limit = 600,
      fastcore_epsilon = 1e-4,
      max_support_reactions = 2000,
      strict = TRUE
    )
  )
)
```

# Sample-aware bootstrap

For condition \(c\), let \(D_c\) be its observed sample/donor identifiers. Each replicate draws \(|D_c|\) sample IDs with replacement. Every selected sample contributes all cells assigned to that sample and condition. This preserves donor clusters; replicate cell counts may vary.

When `sample_col = NULL`, or the named column does not exist, RegCompass prints the exact fallback reason and draws \(n_c\) cells with replacement inside each condition. An existing sample column with missing or empty IDs is rejected.

Both modes re-centre bootstrap target and predictor matrices within condition and refit at the full-data lambda. Selection frequency and sign stability are calculated from the resulting direct condition coefficients.

```{r}
step1 <- readRDS(
  "RegCompass_result/01_single_cell_grn/step_grn.rds"
)
step1$grn_result$bootstrap_policy
head(step1$grn_result$stability_diagnostics[, c(
  "bootstrap_method",
  "bootstrap_resampling_unit",
  "bootstrap_sample_col",
  "bootstrap_fallback_reason",
  "selection_frequency",
  "sign_stability"
)])
```

Sample-cluster bootstrap measures reproducibility across observed samples. Cross-validation remains cell-level, and neither output is a donor-level treatment-effect test.

# Condition genes, reactions, and shared GEM

A gene enters the condition target set when at least one active edge targets it. For reaction \(r\) with GPR branches \(B_{r,k}\):

\[
Core_{r,c}=1
\iff
\exists k:B_{r,k}\subseteq G_c.
\]

Condition core sets may differ. Stage 3 merges their reaction catalogues before Stage 5 constructs one union GEM per medium. All conditions and metacells then reuse identical reaction IDs, stoichiometric matrix, lower bounds, and upper bounds.

# Stepwise inspection

```{r}
step1 <- rc_regcompass_step_grn(
  object = A,
  gem = gem,
  outdir = "RegCompass_steps/01_grn",
  genome = BSgenome.Hsapiens.UCSC.hg38,
  species = "human",
  condition_col = "Group",
  celltype_col = "cell_type",
  sample_col = "sample_id",
  grn_mode = "multitask_shared_backbone",
  pando_args = list(
    pando_design_args = list(peak_to_gene_method = "GREAT")
  )
)

step2 <- rc_regcompass_step_metacells(
  object = A,
  outdir = "RegCompass_steps/02_metacells",
  condition_col = "Group",
  celltype_col = "cell_type",
  fragment_files = FALSE
)

step3 <- rc_regcompass_step_meta_modules(
  grn = step1,
  metacells = step2,
  gem = gem,
  outdir = "RegCompass_steps/03_meta_modules"
)

step4 <- rc_regcompass_step_layer1(
  metacells = step2,
  meta_modules = step3,
  gem = gem,
  outdir = "RegCompass_steps/04_layer1",
  regulatory_alpha = 1,
  gpr_and_method = "min"
)

step5 <- rc_regcompass_step_layer2(
  layer1 = step4,
  meta_modules = step3,
  gem = gem,
  medium_scenarios = medium_scenarios,
  outdir = "RegCompass_steps/05_layer2",
  model_mode = "meta_module_gem",
  layer2_args = list(target_direction = "both", solver = "highs")
)

result <- rc_regcompass_step_results(
  grn = step1,
  metacells = step2,
  meta_modules = step3,
  layer1 = step4,
  layer2 = step5,
  gem = gem,
  outdir = "RegCompass_steps/06_results",
  species = "human"
)
```

Each public stage prints elapsed time and final status in R after its final artifact is committed. Timing is not stored in stage objects, the compact result, or timing TSV files.
